---
title: "When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents"
author: ""
abstract: |
bibliography: "bib/exposure.bib"
csl: apa.csl
output:
  bookdown::pdf_document2:
   # bookdown::word_document2:
    keep_tex: true
    toc: false
    number_sections: false
header-includes:
  - \usepackage{lscape}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
---

\small



**Abstract:**

Can exposure to forgiving partners improve interpersonal cooperation? Attachment theory suggests positive relational experiences can correct negative internal working models, but individuals high in rejection sensitivity (RS), characterized by anxious expectations of rejection, may be resistant to such corrective experiences due to stable negative expectations. We tested this using a randomized experiment (N = 206) in which participants played repeated trust games with HMM-based artificial agents that simulate human-like trust dynamics. After a baseline game, participants were exposed to either forgiving and behaviorally consistent agents (designed to never remain in a low-trust state and free from pre-programmed trust withdrawals; Manipulation) or human-like agents maintaining typical trust withdrawal patterns (Control), then played a final game with a standard agent. Overall, exposure to these idealized partners *reduced* subsequent cooperation; participants appeared to perceive the standard post-exposure agent as less cooperative by comparison, a pattern we interpret primarily as a negative contrast effect. RS moderated specific behavioral patterns but not overall cooperation levels: high RS participants were slower to recover cooperation after a trust withdrawal and became *less* responsive to partner behavior following exposure, whereas low RS participants showed normal recovery and became *more* responsive. These findings suggest that exposure to forgiving partners does not universally promote cooperation, and point to expectation-updating, rather than exposure alone, as a candidate target for future interventions in high RS individuals.

\vspace{1cm}

**Keywords:** Interpersonal functioning; Rejection Sensitivity;
Forgiveness Exposure; Trust-based Cooperation; Hidden Markov Models

<!--
\vspace{1cm}

**General Scientific Summary:**

This study found that exposing people to forgiving partners (with no trust withdrawal) in economic games decreased their subsequent cooperation—likely because human-like partners seemed less cooperative by comparison. Individuals high in rejection sensitivity showed a distinct pattern: while they detected trust withdrawals as readily as others, they were slower to restore cooperation afterward and became less responsive to their partner's behavior. In contrast, those low in rejection sensitivity appeared to learn from the positive exposure and became more reciprocal. These findings suggest that simply providing positive social experiences may not benefit everyone equally, and that individuals prone to rejection sensitivity may need targeted support to translate positive experiences into lasting behavioral change.
-->

\pagebreak

```{r setupCoax, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, include=FALSE) 
knitr::opts_chunk$set(out.width = "\\textwidth")
```

```{r load-packages05}
library(papaja)
library(kableExtra)
require(knitr)
#require(citr)
require(bookdown)

# using some functions dplyr, ggpubr, PairedData and sjPlot. Need to be loaded. 
library(tidyverse)
library(afex)
library(PairedData)
library(multcompView)
library(lsmeans)
library(depmixS4)
library(flextable)
library(grid)
library(gridExtra)
library(forcats)
library(ggsignif)
library(magick)

# Helper function to format emmeans contrast results in APA style
format_emmeans_apa <- function(result_row) {
  est <- result_row$estimate
  se <- result_row$SE
  df <- result_row$df
  t_val <- result_row$t.ratio
  p_val <- result_row$p.value

  # Format p-value
  if (p_val < .001) {
    p_str <- "p < .001"
  } else {
    p_str <- paste0("p = ", papaja::printp(p_val))
  }

  # Format the string
  sprintf("$\\Delta M$ = %.2f, $SE$ = %.2f, $t$(%.1f) = %.2f, %s",
          est, se, df, t_val, p_str)
}

```

```{r analysis-preferences, include = FALSE}
# Seed for random number generation
set.seed(42)
options(tinytex.verbose = TRUE)
```

```{r emmOptions}
emm_options(pbkrtest.limit = 6210)
emm_options(lmerTest.limit = 6210)

```

# Introduction

Trust is fundamental to human social interactions, enabling cooperation
at both the interpersonal and intergroup levels. The study of
psychopathology has linked deficits in trust-based constructs to the
development of mental health disorders [@fonagy_mentalizing_2017].
Individuals with personality disorders (PD) often struggle to form and
maintain social connections, a difficulty reflected in uncooperative
behaviors, a marker for the severity of PD symptoms
[@herpertz_social-cognitive_2014; @mulder_relationship_1999].

One explanation for such social challenges lies in early caregiver
experiences. Attachment theory [@bowlby_attachment_1978] suggests that
the quality of these relationships shapes our capacity for secure
attachments and trust. Individuals with higher levels of insecure
attachment may recall negative trust-related experiences more easily,
report fewer positive trust experiences, and use less constructive
coping strategies when trust is broken [@mikulincer_attachment_1998].
These insecure attachment patterns are often associated with heightened
rejection sensitivity (RS), a tendency to anxiously expect, readily
perceive, and intensely react to rejection [@downey_implications_1996;
@downey_early_1997]. RS has been linked to the development of various
mental health conditions, including depression, anxiety, personality
disorders, and self-harm [@gao_associations_2017]. Individuals high in
RS show attentional biases towards social threat cues, which may
contribute to difficulties in social interactions
[@berenson_rejection_2009]. A recent review highlighted prosocial
behavior and interpersonal trust as two key processes of interpersonal
functioning that are markedly impaired in PDs and which are likely to
contribute to interpersonal dysfunction in this population
[@hepp_prosociality_2022]. The interaction of RS and trust-based
constructs has been explored, particularly in Borderline Personality
Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found
that RS mediated the relationship between BPD features and lower trust
appraisal. @abramov_influence_2022-1 found that higher baseline feelings
of rejection in individuals with BPD predict slower trust formation and
less pronounced declines in trust following trust withdrawals during the
trust game. However, the interaction between *reciprocity* and RS hasn't
been studied as extensively, leaving a gap in our understanding of how
these constructs might interplay.

Given that RS may be a manifestation of maladaptive attachment styles,
it is important to explore whether exposure to consistently forgiving
and reliable interaction partners could reshape interpersonal expectations
and behaviors. The *corrective experience hypothesis*, rooted in attachment
theory, suggests that new positive relational experiences can modify
internal working models of relationships [@bowlby_secure_1988]. Research
on social learning [@bandura_social_1977] similarly demonstrates that
individuals model the behavior of those around them, and exposure to
cooperative peers promotes cooperative behavior
[@fowler_cooperative_2010]. In the repeated trust game (RTG) paradigm,
cycles of reciprocated trust enhance cooperative behaviors even among
initially distrustful individuals [@king-casas_getting_2005]. This
perspective predicts that exposure to forgiving partners should increase
subsequent cooperation, as participants internalize more positive
expectations about social interactions.

However, an alternative perspective suggests high RS individuals may be
*resistant* to such corrective experiences. RS is characterized by stable
negative expectations that operate through self-fulfilling prophecies
[@downey_implications_1996]: high RS individuals interpret ambiguous social
cues negatively, which elicits rejection, thereby confirming their
expectations. Research on belief updating in depression and personality
pathology has documented "cognitive immunization" processes whereby negative
schemas resist modification despite contradictory evidence [@kube_distorted_2020].
From this perspective, positive exposure might fail to update expectations
in high RS individuals, or might even produce paradoxical effects if the
contrast between positive exposure and subsequent "normal" interactions
confirms their belief that trustworthy partners are rare.

In this study, we use a randomized controlled online experiment to test
whether exposing participants with varying RS levels to forgiving and
more cooperative co-players results in more trustworthy behavior and a
repair of potential breakdowns in RTG cooperation. To simulate realistic
social interaction while maintaining a high degree of experimental
control, we take a novel paradigmatic approach: We use generative models
of how humans play the RTG to design an agent that plays the role of the
investor, based on Hidden Markov Models (HMMs) fitted to real players'
data. A key aspect of these agents is that their actions depend on a
latent "trust state" which reacts dynamically to the trustees' returns,
simulating real-life trust-building scenarios. An advantage of having
such a generative model of behavior is the possibility of controlling
different aspects of the agent's strategy such as its general policy,
the propensity to cooperate actively, or the propensity to trust again
after breakdowns of cooperation. To further mimic real-world
interactions and examine participants' responses to one-off breakdowns
of cooperation, we incorporate occasional pre-programmed low investments
by the agent.

We pre-screened participants for high or low RS using a validated
questionnaire, then assigned them exclusively to the trustee role in a
series of trust games. After playing a 15-round RTG with a human-like
HMM investor, they were randomly assigned to either a Control or
Manipulation condition. In the Manipulation condition, participants were
exposed over three RTGs to HMM investors designed with a limited
propensity for retaliation: agents that were both forgiving of low returns
*and* free from pre-programmed trust withdrawals, with this pattern
maintained across all three games of the exposure phase. In the Control condition, participants
played three RTGs against the same human-like HMM that maintained the
occasional low-investment pattern from the pre-exposure phase,
representing continuity with typical social interactions. This design
tests whether exposure to partners combining forgiveness with behavioral
consistency transfers to subsequent interactions with standard partners. After this exposure phase, all participants played
another 15-round RTG with a human-like HMM investor, similar to the one
in the pre-exposure phase.

Based on the corrective experience hypothesis, we predicted that
exposure to these idealized, forgiving partners would increase subsequent
cooperation, with high RS individuals potentially benefiting from positive
relational experiences that challenge their negative expectations. Because
the Manipulation condition combined forgiveness with behavioral consistency,
our design tests a strong version of this hypothesis: whether sustained
exposure to forgiving and behaviorally consistent partner behavior
transfers to subsequent interactions. We examined both overall effects of the manipulation and
differential responses based on RS, with particular attention to how
participants respond to trust withdrawals before and after the exposure
phase. We note that while our primary prediction was grounded in the
corrective experience hypothesis, the alternative interpretations
involving negative contrast effects and RS-specific updating deficits
were refined following initial data analysis (see Transparency and Data
Availability) and should be considered exploratory. The moderation of
the manipulation effect by RS was itself a planned analysis, reflected
in the extreme-groups design, though it was not separately powered (see
Participants).

# Methods

```{r loadData}

load("data/d_finished.RData")

# Step 1: Identify the expo_opponent for each playerId where gameNumber == "11" (exposure game) & isLastRound == TRUE (to select a particular)
expo_opponent_df <- d_finished %>%
  filter(gameNumber == "11", isLastRound == TRUE) %>%
  dplyr::select(id, expo_opponent = game_opponent) %>%
  distinct() # Ensure uniqueness in case of duplicates

# Step 2: Join this information back to the original dataset
d_finished <- d_finished %>%
  left_join(expo_opponent_df, by = "id") %>%
  mutate(condition.f = factor(expo_opponent, levels= c("AI_HMM","AI_HMM_nice"),
                              labels=c("Control","Manipulation"))) %>%
  dplyr::select(-expo_opponent) # remove the expo_opponent column if it's no longer needed


# Step 3:  create binary : high or low RS
d_finished <- d_finished %>%
  mutate(high_RS = factor(ifelse(RS_score > 15, 1, 0), 
                                    levels = c(0, 1), 
                                    labels = c("low RS", "high RS"))) 


# Summarize to count the number of participants for each Condition within each high_RS level
temp <- d_finished %>% dplyr::select(id, condition.f, high_RS) %>% 
                      unique() %>%
                      group_by(high_RS, condition.f) %>%
                      summarize(participant_count = n(), .groups = 'drop')

# View the summarized counts
print(temp)

###################### Extract latent Investor State in each round ####################

# Select the desired columns
d_selected <- d_finished %>%
  dplyr::select(roundNum, gameNumber, playerId, starts_with("State_"))

# Reshape the data
d_reshaped <- d_selected %>%
  pivot_longer(cols = starts_with("State_"), 
               names_to = "State_name", 
               values_to = "investorState") %>%
  mutate(roundNum = as.numeric(str_extract(State_name, "(?<=r)\\d+")),
         gameNumber = as.numeric(str_extract(State_name, "(?<=game_)\\d+"))) %>%
  dplyr::select(playerId, gameNumber, roundNum, investorState) %>%
  arrange(playerId, gameNumber, roundNum) %>%
  filter(!is.na(investorState)) %>% unique()

# Join the reshaped data back to the original dataset
d_finished <- d_finished %>%
  dplyr::select(-investorState) %>%
  left_join(d_reshaped, by = c("playerId", "gameNumber", "roundNum"))
```

```{r bonusPayment, include=FALSE}
# -----------------------------------------------------------------------------
# Realised performance-contingent bonus from game-unit trust payoffs.
# Documents the conversion from game units to GBP (Reviewer 2, comment 8).
# Placed here so the realised-bonus values can be reported once, in the
# Participants compensation sentence below, avoiding duplication.
# -----------------------------------------------------------------------------

base_fee_gbp         <- 6.00
target_avg_bonus_gbp <- 0.50

# Total game-unit payoff per participant, summing all five trust games.
trust_payoffs <- d_finished %>%
  dplyr::distinct(playerId, .keep_all = TRUE) %>%
  dplyr::mutate(total_trust_units = payoffTrust1 + payoffTrust2 +
                                    payoffTrust11 + payoffTrust12 + payoffTrust13) %>%
  dplyr::select(playerId, total_trust_units) %>%
  dplyr::filter(!is.na(total_trust_units))

mean_trust_units <- mean(trust_payoffs$total_trust_units)

# Conversion calibrated to deliver the target average bonus.
gbp_per_unit <- target_avg_bonus_gbp / mean_trust_units

# Per-participant realised GBP bonus.
realised_bonuses_gbp <- trust_payoffs$total_trust_units * gbp_per_unit
mean_bonus_gbp <- mean(realised_bonuses_gbp)   # = target by construction
sd_bonus_gbp   <- sd(realised_bonuses_gbp)
min_bonus_gbp  <- min(realised_bonuses_gbp)
max_bonus_gbp  <- max(realised_bonuses_gbp)
bonus_to_fee_ratio_denom <- round(base_fee_gbp / mean_bonus_gbp)
```

## Participants

To have participants with large differences in RS, a total of 1195
participants were pre-screened on the Prolific Academic platform
(prolific.co) using the Rejection Sensitivity Questionnaire (RSQ) to
finally select two similarly sized groups: One with high RS (RSQ score
\> 15, N=103) and the other with low RS (RSQ score \< 10, N=103)
totalling 206 participants (56% female). These were then invited through
prolific to take part in the main experiment. The required sample size
was determined using an *a priori* power analysis to have an 80%
probability to detect a small effect size (Cohen’s f = 0.10) for a
within-between interaction with a 5% type I error rate in a repeated
measures ANOVA. The sample size calculation assumed two groups, two
measurements per group and was performed using the G\*Power software
[@faul_statistical_2009]. This power analysis targeted the Condition
$\times$ Phase interaction (the effect of the manipulation on returns);
moderation by RS, although central to the design through the
extreme-groups selection described above, was not separately powered,
and RS-related analyses are therefore interpreted with corresponding
caution. The mean age of participants was 34.6 years,
with an 11.9 years standard deviation. The majority of participants
identified ethnically as White ($80$%). The online cohort registered 30
unique countries of birth with the most frequent being the U.K ($33$%)
followed by Poland ($10$%) and Portugal ($10$%). Participants were paid
a fixed fee of £`r format(base_fee_gbp, nsmall = 2)` plus a
performance-contingent bonus computed from their total points across
all rounds of the five trust games, calibrated so that the realized bonus averaged
£`r format(round(mean_bonus_gbp, 2), nsmall = 2)` per participant
(SD = £`r format(round(sd_bonus_gbp, 2), nsmall = 2)`; range
£`r format(round(min_bonus_gbp, 2), nsmall = 2)`–£`r format(round(max_bonus_gbp, 2), nsmall = 2)`),
approximately one `r ifelse(bonus_to_fee_ratio_denom == 12, "twelfth", paste0(bonus_to_fee_ratio_denom, "th"))`
of the fixed fee. Participants were informed at the outset that the
bonus would be proportional to the points accumulated in the games; the
exact conversion rate was not communicated in advance. Data were
collected over multiple sessions.

## Design and procedure

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS :
High or Low) by 2 (Phase: Trust-Game Pre-Exposure, Trust-Game Post-Exposure) design, with repeated measures on the Phase factor (Figure
\@ref(fig:HMMPanels).A). Participants within each pre-screened RS group
were randomly assigned to one of the two levels of the Condition factor,
resulting in 101 participants in the Manipulation condition and 105 in
the Control condition. The games were designed and implemented online
using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment
received approval from the university's
ethics commission (ID:S-708/2023) and the experiment was performed in

accordance with the ethics board guidelines and regulations. All
participants provided informed consent prior to their participation.

## Tasks and measures

```{r, include=FALSE}
Anti_social <-  c(0.1025436430408, 0.1221931236853, 0.1345215238995,
0.1368182252617, 0.1285592471751, 0.1116014322677, 0.0895041220882,
0.0663166721334, 0.0453950277118, 0.0287077419587, 0.0167723588011,
0.0090530028158, 0.0045143317507, 0.0020796744990, 0.0008851094702,
0.0003480141146, 0.0001264134789, 0.0000424213859, 0.0000131513231,
0.0000037665630, 0.0000009965755)

Neutral <- c(0.003393529, 0.006930874, 0.013053553, 0.022671228,
0.036310144, 0.053627606, 0.073039389, 0.091734929, 0.106248217,
0.113479701, 0.111769796, 0.101517390, 0.085028780, 0.065675083,
0.046778251, 0.030725245, 0.018610323, 0.010394861, 0.005354126,
0.002543094, 0.001113881)

Pro_social <- c(0.003162697, 0.001057162, 0.001410197, 0.001881016,
0.002508877, 0.003346113, 0.004462480, 0.005950950, 0.007935434,
0.010581067, 0.014107909, 0.018809194, 0.025075644, 0.033427845,
0.044559370, 0.059394206, 0.079163228, 0.105506029, 0.140606514,
0.187373417, 0.249680651)

investment <- seq(0:20) -1

response_probs <- as.data.frame(cbind(investment,Anti_social,Neutral,Pro_social)) %>% 
  rename("low-trust" = "Anti_social", "medium-trust"="Neutral", "high-trust" = "Pro_social") %>%
  pivot_longer(cols=c("low-trust","medium-trust","high-trust"),
                    names_to='Investor_state',
                    values_to='probability') %>% 
   mutate(across(Investor_state, factor, levels=c("low-trust","medium-trust","high-trust")))
```

```{r, include=F}

plotinvHMM <- ggplot(response_probs,                            
       aes(x = investment,
           y = probability,
           fill = Investor_state)) +
  geom_bar(stat = "identity",
           position = "dodge") + 
  labs(fill='Latent investor state', x = "Investment", y= "Probability") + 
  theme_bw() + 
  theme(legend.position = "bottom",  legend.text = element_text(size = 12),  legend.title = element_text(size = 14))
  
plotinvHMM
```

```{r, include=FALSE}

# Parameters for HMM human-like Transition Function
unhappy_pars <- rbind(c(0,0), c(-3.366027, 0.40910797 ), c(-3.572619,0.08137274)) 
neutral_pars <- rbind(c(0,0), c(3.3142637, 0.3763408), c(0.9169736, 0.4502838))
happy_pars   <- rbind(c(0,0), c(0.7134085, 0.02101626), c(2.2215478 ,0.16162964))

pars_inv <- list(unhappy_pars, neutral_pars, happy_pars)


# Parameters for HMM nice Transition Function
unhappy_pars_nice <- rbind(c(-10,0), c(-3.366027, 0.40910797 ), c(-3.572619,0.08137274)) 
neutral_pars_nice <- rbind(c(0,0), c(3.3142637, 0.3763408), c(0.9169736, 0.4502838))
happy_pars_nice   <- rbind(c(0,0), c(0.7134085, 0.02101626), c(2.2215478 ,0.16162964))

pars_inv_nice <- list(unhappy_pars_nice, neutral_pars_nice, happy_pars_nice)



plot_HMM_transitions <- function(ns, pars_mat) {

  trans_prob <- data.frame(
    from = rep(1:ns, each=100*ns),
    to = rep(1:ns, each=100),
    ret = seq(-20,60,length=100),
    probs = 0
  )
  
  
  y <- matrix(0.0,ncol=ns, nrow=100)
  
  for(from in 1:ns) {
  pars <- matrix(pars_mat[[from]], ncol=2)
  # print(pars)
  
    for(to in 1:ns) {
        x <- trans_prob[trans_prob$from == from & trans_prob$to == to,"ret"]
        y[,to] <- exp(pars[to,1] + pars[to,2]*x)
    }
    y <- y/rowSums(y)

    
    for(to in 1:ns) {
      trans_prob$probs[trans_prob$from == from & trans_prob$to == to] <- y[,to]
    }
  }
  
  df <- as.data.frame(trans_prob) %>% 
    mutate(from = recode(from, "1" = "low-trust", "2" = "medium-trust", "3" = "high-trust"),
           to = recode(to, "1" = "low-trust", "2" = "medium-trust", "3" = "high-trust") ) %>% 
    mutate(across(from, factor, levels=c("low-trust","medium-trust","high-trust"))) %>% 
    mutate(across(to, factor, levels=c("low-trust","medium-trust","high-trust")))
                                    
  
    # Create a separate data frame with the background colors
  bg_colors <- data.frame(
    from = factor(c("low-trust", "medium-trust", "high-trust"), levels=c("low-trust","medium-trust","high-trust"))
  )
  
  # plotting code...
  ggplot() +
    geom_rect(data = bg_colors, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf, fill = from), alpha = 0.1) +
    geom_line(data = df, aes(x = ret, y = probs, colour = as.factor(to))) +
    facet_wrap(~from, labeller = labeller(from = function(x) paste("From", x, "state on trial t"))) +
    ylim(c(0,1)) +
    scale_fill_manual(values = c("low-trust" = "red", "medium-trust" = "green", "high-trust" = "blue"),
                    name = "From state") +  # Changed legend title for 'fill' here
    scale_color_manual(values = c("low-trust" = "red", "medium-trust" = "green", "high-trust" = "blue"),
                       labels = c("low-trust", "medium-trust", "high-trust"),
                       name = "State transitioned to") +
    labs(x = "Investor's net return on trial t", y = "Transition probability to \nState on trial t+1", color = 'State transitioned to') +
    theme_bw() +
    theme(legend.position = "bottom",
          legend.text = element_text(size = 12),
          legend.key.size = unit(1, 'lines'),
          legend.spacing.x = unit(0.1, 'in'),
          legend.title = element_text(size = 14),
          legend.margin = margin(t = 0.2, b = 0, unit = 'cm'),
          plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "cm"),
          strip.text = element_text(size = 10),
          legend.box = "vertical" # Arrange legends vertically
    ) +
    guides(fill = guide_legend(order = 1, title.position = "left", title.hjust = 0.3),
           color = guide_legend(order = 2, title.position = "left", title.hjust = 0.3))
}

```

```{r, include=F}

plotInvTran <- plot_HMM_transitions(3, pars_inv) 
print(plotInvTran)

plotInvTranNice <- plot_HMM_transitions(3, pars_inv_nice) 
print(plotInvTranNice)


```

### Repeated trust game and HMM investor

Participants played a 15-round RTG [@joyce_trust_1995] in the trustee
role against a computer-programmed investor. We adopted a
fixed-trustee design for three reasons. Our research question concerns
reciprocity (the trustee's response to a partner's investment), which
is a distinct construct from trust propensity [@cox_what_2016]. The
HMM was fitted as a generative model of investor behavior only.
Fixed-role designs are also standard in the iterated trust-game
literature to avoid role-switching confounds [@joyce_trust_1995;
@king-casas_getting_2005; @king-casas_rupture_2008; @guennouni_price_2026].
On each round the investor
is endowed with 20 units and decides how much of that endowment to
invest. This investment is tripled and the trustee then decides how to
split this tripled amount between them and the investor. If the trustee
returns more than one third of the amount, the investor makes a gain.
Each player was represented with an icon with the participant always on
the left of the screen and the co-player on the right. The participants
were able to choose the icon that represents them at the start of the
experiment. The icon representing the co-player changed at the start of
each new game, to simulate a new interaction partner. Participants were
not told they were facing computerized co-players. We chose to simulate
the behavior of a human interaction partner through allowing for a delay
while pairing with new opponents at the start of each game as well as
programming the agents to respond during each round after a varying time
lapse (randomly chosen between 5 and 10 seconds).

The computerized investor consisted of a hidden Markov model (HMM)
trained on an independent existing behavioral RTG data set of human
investors. This data-driven approach thus sought to learn an investor
strategy that mimics human-like interactions. The data set used for
training consists of 381 ten round games with the same player (full
details can be found in the Supplementary Information). Importantly, the
HMM was fitted to the behavior of healthy investors, while the trustees
in the training data included both healthy participants and individuals
with personality disorders, providing a diverse range of interaction
dynamics from which the investor strategy was learned.

We retained this mixed-trustee training set deliberately. Because the HMM models *investor* behavior, and the investors were healthy, the trustee heterogeneity enriches only the range of partner behavior the agent learned to handle, rather than contaminating the agent itself. The wider, more variable returns produced by personality-disorder trustees in fact better identify the investor's transitions out of low trust, and better match our high-RS test sample, given that high rejection sensitivity is commonly associated with borderline personality features [@miano_rejection_2013; @richetin_emotional_2018].

On this data set, the HMM was inferred with three latent states that could be
interpreted as reflecting a “low-trust”, a “medium-trust”, and a
“high-trust” state. A separate output distribution, that maps each HMM
state onto possible investments from 0 to 20 separately, is learned
(Figure \@ref(fig:HMMPanels).B). In analogy to the latent states, these
distributions can be interpreted as reflecting “low-trust”,
“medium-trust”, or “high-trust” dispositions. Finally, the HMM is
specified by transition probabilities that describe the transition
between states. The probability of these transitions was modeled as a
function of their net return (i.e return - investment) in the previous
round (see Figure \@ref(fig:HMMPanels).C). The initial state for the
HMM investor in each instance of the game was set to the "medium-trust"
state. Details on how the HMM state conditional probabilities and
transition functions are specified can be found in the Supplementary Information (Section B).

In order to instigate a potential breakdown of trust, thereby allowing
us to probe efforts to repair it, the computerized agent was programmed
to provide a low investment on round 12 (pre-exposure) and round 13
(post-exposure). On all other rounds, the investor’s actions were
determined by randomly drawing an investment from the state-conditional
distribution, with the state over rounds determined by randomly drawing
the next state from the state-transition distribution as determined from
the net return on the previous round (disregarding the net return
immediately after the pre-programmed low investment rounds).

## Manipulation

In all phases of the RTG other than the ‘Exposure phase’ (Figure
\@ref(fig:HMMPanels).A), participants interacted with this human-like
HMM. In the ‘Manipulation’ Condition of the exposure phase, however, the
parameters of this HMM were adjusted to design a ‘forgiving’ and
ultimately more cooperative agent. To achieve this, we changed the state
transition probabilities of the HMM such that it becomes impossible for
it to remain in a low trust state, effectively setting the transition
probability for remaining in a “low-trust” state to 0. The resulting
transition function is shown in Figure \@ref(fig:HMMPanels).D. The
policies conditional on the latent states and the transition function in
the other latent states remain unchanged. Critically, the transitions
*into* the low-trust state were left unchanged: the forgiving agent
still responds to a low return by reducing its next investment, exactly
as the standard agent does. It differs only in that it cannot dwell in
low trust; it recovers to a higher-trust state by the following round.
The forgiving agent is thus equally contingent on the participant's
returns in the short term, differing from the standard agent in the
*persistence* rather than the *presence* of its response to low returns.
The Manipulation condition
therefore combines two design features: forgiveness (immediate exit from
the low-trust state) and the absence of programmed trust withdrawals
during exposure (in contrast to the Control condition, which
additionally delivered a low investment in round 5 of each exposure
game; see Procedure). The experimental contrast tests their joint effect
rather than isolating forgiveness alone. Because the forgiving agent
could not dwell in the low-trust state, its realized investments during
exposure were also higher on average; this is a downstream consequence
of the altered transition dynamics rather than a separately manipulated
feature.

## Procedure

At the start of the experiment, participants provided informed consent
and were instructed the study would consist of three phases in which
they would face a different other player. Participants were told their
goal was to maximize the number of points in all phases. They were not
told the number of rounds of each phase. Participants were randomly
assigned to either a Control or Manipulation condition. The timeline of
the experiment is shown in Figure \@ref(fig:HMMPanels).A. Phase one
("pre") consisted of a 15 round RTG in which participants took the role
of trustee, facing the same investor over all 15 rounds. On each round,
after being informed about the amount sent by the investor participants
decided how much of the tripled investment to return to the investor,
before continuing to the next round. Phase 2 ("exposure") consisted of
three seven-round RTGs. Participants in the Manipulation condition faced the
forgiving HMM investor and rated the agent on the same attributes as in
the pre-exposure phase. Those in the Control condition faced the
same human-like HMM agent as in the "pre" phase and rated each co-player
on the same attributes. To keep the experience similar to the "pre"
phase, the agent in the Control condition was also designed to send a
very low investment in round 5 of each of the three games. In the
post-exposure phase ("post"), participants in both conditions faced
the same human-like HMM as in "pre" phase.

At the beginning of each game in all three phases, participants were
explicitly told they would face a **new** player, and saw a simulated
pairing-with-an-available-co-player wait screen that mimicked the
connection delay of online multiplayer experiments. Combined with the
change of the co-player icon at each new game (described above), this
framing was designed to discourage participants from treating
cross-game behavior as continuous with a single partner. After completing each RTG in each phase, participants
rated how cooperative and forgiving they perceived the co-player to be,
and whether they would like to play with them again (all on a scale from
1 to 10 with 10 being the most positive rating). After completing the
three game phases, participants then completed the Levels of Personality
Functioning Scale Brief-Form (LPFS-BF) questionnaire
[@weekers_level_2019]. This is a self-report measure designed to assess
core elements of personality functioning as defined in the Alternative
Model for Personality Disorders in the DSM-5
[@american_psychiatric_association_diagnostic_2013], and provides a
dimensional assessment of personality functioning, which complements the
categorical approach of RS. Finally, participants were asked whether
they thought the other players were human or computer agents, to probe
how well the agent can mimic human behavior, then debriefed and thanked
for their participation.

````{=tex}
\begin{landscape}
```{r HMMPanels, fig.cap= "A: Experiment timeline. Participants (trustees) played RTGs with HMM investor agents. The investor sends investments (multiplied by 3) and participants decide returns. Conditions differ in exposure phase agents. B-D: The artificial investor is a three-state HMM fitted to human data. B: Investment distributions by latent state. C: Transition probabilities to states at t+1 as a function of net return at t; each panel shows a different starting state. D: Forgiving HMM transitions from low-trust state: unlike the human-like HMM, it always exits low-trust and favors high-trust transitions.", include=T, out.width='0.95\\linewidth', fig.align='center'}


library(magick)

# Read in the images
image1 <- image_read("figures/timeline.png")
image2 <- image_read("figures/HMMinvPolicy2.png")
image3 <- image_read("figures/HMMinvTrans2.png")
image4 <- image_read("figures/HMM_switch.png") 



# Determine the target width for all images, for example, the width of the widest image
target_width <- max(image_info(image1)$width, image_info(image2)$width, 
                    image_info(image3)$width, image_info(image4)$width)

target_height <- max(image_info(image1)$height, image_info(image2)$height, 
                    image_info(image3)$height, image_info(image4)$height)


# Scale images to have the same width
image1 <- image_scale(image1, paste0(target_width, "x", target_height))
image2 <- image_scale(image2, paste0(target_width, "x", target_height))
image3 <- image_scale(image3, paste0(target_width, "x", target_height))
image4 <- image_scale(image4, paste0(target_width, "x", target_height))

# Annotate each image with a label
image1 <- image_annotate(image1, text = "A", location = "+0+10", size = 90, color = "black", gravity = "northwest")
image2 <- image_annotate(image2, text = "B", location = "+0+10", size = 90, color = "black", gravity = "northwest")
image3 <- image_annotate(image3, text = "C", location = "-0+10", size = 90, color = "black", gravity = "northwest")
image4 <- image_annotate(image4, text = "D", location = "+0+10", size = 90, color = "black", gravity = "northwest")

# Assuming a 300 DPI resolution for good print quality
dpi <- 300

# Spacing of 0.1 inches converted to pixels
spacing <- 0.05 * dpi

# Calculating the width and height for each image (with shorter caption, can use more space)
width_per_image <- ((10.5 - 0.1) / 2) * dpi
height_per_image <- ((5.8 - 0.1) / 2) * dpi
# The geometry string for each image, including spacing (assuming no borders)
geometry_string <- paste0(width_per_image, 'x', height_per_image, '+', spacing, '+', spacing)

# Now create the montage with the specified geometry
combined_image <- image_montage(
  c(image1, image2, image3, image4), 
  tile = "2x2",
  geometry = geometry_string
)


# Display the combined image
plot(combined_image)
```
\end{landscape}
````

## Statistical analysis

We analyzed participants' behavior in the RTG using linear mixed-effects models. First, to examine the effect of the manipulation, we modeled the percentage return (percentage of tripled investment returned to investor) as a function of Phase (RTG game pre vs. post-exposure), Condition (Manipulation vs. Control), Investment, and RS (High vs Low RS group), including all interactions as fixed effects. This model included player-wise random intercepts and slopes for Phase. Second, we analyzed behavior during the Exposure phase specifically, modeling returns with Condition, Investment, and RS and their interactions as fixed effects, along with player-wise random intercepts. Third, to verify the consistency of the HMM agent, we modeled the investments sent by the computerized agent using Condition, Phase, and RS and their interactions as fixed effects. To isolate effects occurring prior to any pre-programmed low investment, we also analyzed returns in rounds preceding the low investment trials only (rounds 1-11 in the pre-exposure phase and rounds 1-12 in the post-exposure phase) using the same model specification. To test whether reduced returns reflected strategic exploitation, we examined investor-harming returns (those below one-third of the tripled investment, which cause the investor to incur a net loss) using a mixed logistic regression with Phase, Condition, Investment, and RS as predictors, and player-wise random intercepts. Trustee payoffs were compared between phases using paired t-tests, and between conditions in the post-exposure phase using a Welch t-test and a Mann-Whitney test. Finally, to rigorously assess participants' reactions to and recovery from the specific instance of pre-programmed low investment, we conducted an event study analysis centered on the low investment round ($t=0$). We fit a linear mixed-effects model to percentage returns in the rounds surrounding the low investment ($t-2$ to $t+2$), with Phase, Condition, Time Point, and RS Group as fixed effects. We examined the Drop (change in return from $t-1$ to the low investment round) and the Recovery (change in return from the low investment round to $t+1$), and tested whether the between-condition difference in Recovery persisted at $t+2$. To assess sensitivity to suspicion about the agent, we additionally modeled returns with Phase, Condition, and detection status (whether the participant identified the agent as a machine in the post-experiment question) as factors, testing whether the manipulation effect depended on detection. We modeled the exposure-phase player ratings as a function of Condition and RS as a manipulation check. The specification of the main statistical models and their coefficient tables can be found in the supplement.

All models were estimated using the `afex` package [@singmann_afex_2022] in R. We determined the random effects structure by starting with the maximal model and simplifying until convergence was achieved, ensuring the optimal structure [@matuschek_balancing_2017]. A similar process was applied to the models analyzing HMM agent investments and participant ratings. We report differences in marginal means rather than standardized effect sizes for the focal contrasts, as there is no consensus on effect size calculation for mixed models. As a descriptive summary of model fit, we also report, for each headline model, marginal $R^2$ (variance attributable to the fixed effects) and conditional $R^2$ (variance attributable to both fixed and random effects), computed with the `performance` package (version `r as.character(packageVersion("performance"))`) following Nakagawa and Schielzeth [-@nakagawa_general_2013] and Nakagawa, Johnson, and Schielzeth [-@nakagawa_coefficient_2017]. These values are reported alongside each model in the Results. $F$-tests used the Kenward-Roger approximation for degrees of freedom, which produces fractional degrees of freedom that vary across effects depending on whether they are between-subjects (lower df) or within-subjects (larger df). The Investment variable was Z-transformed to facilitate the interpretation of main effects in the presence of interactions. Significant interactions were probed using planned contrasts with the `emmeans` package. We applied the "Sidak" correction for multiple comparisons to control the familywise error rate while maintaining statistical power.

```{r datRatings}

#######################    Pre-processing player ratings 

# Pivot longer ratings on each attribute 
df_coop <- d_finished %>% dplyr::select(playerId,condition.f,contains("cooperative"),high_RS) %>%  
  pivot_longer(cols=contains("cooperative"), names_to = c("game"), names_pattern ="rating_cooperative_(.*)", values_to = "rating_coop") %>% distinct()

df_forgiv <- d_finished %>% dplyr::select(playerId,condition.f,contains("forgiving"),high_RS) %>%
  pivot_longer(cols=contains("forgiving"), names_to = c("game"), names_pattern ="rating_forgiving_(.*)", values_to = "rating_forgiving") %>% distinct()

df_playAgain <- d_finished %>% dplyr::select(playerId,condition.f,contains("again"),high_RS) %>%  
  pivot_longer(cols=contains("again"), names_to = c("game"), names_pattern ="rating_playAgain_(.*)", values_to = "rating_playAgain") %>% distinct()

#merge all data frames together
datRatings <- list(df_coop, df_forgiv,df_playAgain) %>% 
              reduce(full_join, by=c('playerId','game','condition.f','high_RS')) %>%
              mutate(gameNum.f = factor(game,labels = c("pre", "expo1", "expo2", "expo3","post"),
                                      levels=c("1", "11", "12", "13", "2")))

# Group means for each rating 
mu <- datRatings %>% group_by(gameNum.f, condition.f) %>% summarise(mean_coop = mean(rating_coop),
                                                  mean_forgiv = mean(rating_forgiving),
                                                  mean_playAgain = mean(rating_playAgain)
                                                  )
mu


```

```{r}

df_long_ratings <- datRatings %>%
  pivot_longer(
    cols = starts_with("rating_"), 
    names_to = "rating_type", 
    values_to = "rating_value"
  ) %>%
  dplyr::select(playerId, condition.f, gameNum.f, rating_type, rating_value, high_RS)


# First, calculate the mean and SEM for each group
summary_df <- df_long_ratings %>%
  group_by(gameNum.f, condition.f, rating_type, high_RS) %>%
  summarise(
    mean = mean(rating_value, na.rm = TRUE),
    sem = sd(rating_value, na.rm = TRUE)/sqrt(n())
  ) %>%
  ungroup()


```

<!-- ## Linear mixed effects Model for returns  -->

```{r avgRetDf}
# Filter only trust rounds and create % returns and investments 

avg_ret_df <- d_finished %>%
  dplyr::select(playerId,roundType,investment,returns,roundNum,gameNum.f,condition.f, phase.f, high_RS, LPFS_score, RS_score, investorState.f) %>% 
  filter(roundType=="trust",!is.na(roundNum)) %>% 
  mutate(roundNum = as.numeric(as.character(roundNum))) %>%
  mutate(inv_scaled = as.vector(scale(investment))) %>% 
  mutate(inv_pct = investment/20, ret_pct_0 = ifelse(investment==0,0,returns/(3*investment)),ret_pct_na = ifelse(investment==0,NA,returns/(3*investment))) 

# Number of people left for analysis. (there are 51 rounds in each game)
#num_participants <- nrow(avg_ret_df)/51
num_participants <- length(unique(avg_ret_df$playerId))

num_participants 

```

```{r}
# Number of participants in manipulation condition
temp_manip <- avg_ret_df %>% filter(condition.f =="Manipulation")

length(unique(temp_manip$playerId))

# Number of participants with high RS

temp_RS <- avg_ret_df %>% filter(high_RS =="high RS")
length(unique(temp_RS$playerId))

```

```{r}
# average investment for first 10 rounds pre-exposure
avg_pct <- avg_ret_df %>% filter(roundNum < 11 , gameNum.f == "pre") %>% summarise(avg_inv = mean(inv_pct), avg_ret = mean(ret_pct_na, na.rm=TRUE))
avg_pct
```

<!-- Linear mixed effects model for all returns  -->

# Behavioral Results

## Analysis of participant returns

```{r modAllReturns, include=FALSE, cache=TRUE}
##############  Filter only pre and post games
prepost_df <- avg_ret_df %>% filter(gameNum.f %in% c("pre","post"))
##############
mod_returns_pct_RS <- mixed( ret_pct_na ~ phase.f*condition.f*inv_scaled*high_RS + (1 + phase.f| playerId), prepost_df, REML= TRUE, method="KR")

summary(mod_returns_pct_RS)

```

```{r}
saveRDS(summary(mod_returns_pct_RS), "data/mod_returns_pct_RS.RDS")
```

```{r}
contrasts(prepost_df$phase.f)
contrasts(prepost_df$condition.f)
contrasts(prepost_df$high_RS)
contrasts(prepost_df$investorState.f)
```

```{r prepostByCond, cache=TRUE}
prepost_bycond <- pairs(emmeans::emmeans(mod_returns_pct_RS, c("phase.f"), by = "condition.f", pbkrtest.limit = 6300))
prepost_bycond
```

```{r modemmeans, cache=TRUE, include=F}
library(emmeans)

# Compute EMMs for the interaction of interest
emms <- emmeans(mod_returns_pct_RS, ~ phase.f * condition.f | high_RS, pbkrtest.limit = 6300, adjust="tukey")

contrasts <- contrast(emms, "pairwise", by = c("condition.f", "high_RS"))
summary(contrasts)

results_df <- as.data.frame(summary(contrasts))
```

<!-- EFFECT SIZES -->

```{r}
# Load the necessary packages

library(sjstats)
effectsize::eta_squared(mod_returns_pct_RS)
```

On average, investments and returns, as shown in Figure
\@ref(fig:gamesPlot), fell within the documented range of 40-60% of the
endowment for investments and 35-50% of the total yield for returns, as
reported in previous studies [@charness_investment_2008;
@fiedler_social_2011].

```{r gamesPlot, include=T, echo=FALSE, fig.cap="Averages and standard errors of the trustee's return as a percentage of the multiplied investment received (y-axis) by Condition, Phase, and game round (x-axis) averaged across RS groups. The blue line shows the returns in the Pre phase and the green line those in the Post phase. The left Panel shows returns in the Control condition and the right one those in the Manipulation condition. The dotted lines identify the rounds where the pre-programmed one-off low investment occurs. We note lower average returns post vs pre in the Manipulation condition, while returns in the Control condition are similar between the two phases.",fig.align="center", fig.width=5, fig.height = 3.3}

# Create a data frame for the vertical lines
vline_data <- data.frame(
  xintercept = c(12, 13),
  Low_inv_round = factor(c("Pre-phase", "Post-phase"), levels = c("Pre-phase", "Post-phase"))
)


# Plot
ggplot(avg_ret_df %>% filter(gameNum.f %in% c("pre","post")), aes(x=as.factor(roundNum), y=ret_pct_na, group=gameNum.f, color = gameNum.f, fill=gameNum.f)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(ymin=..ymin.., ymax=..ymax..),
               alpha = 0.3, linetype = 0) +
  geom_vline(data = vline_data, aes(xintercept = xintercept, linetype = Low_inv_round), color = "black", linewidth = 0.6) +
  scale_linetype_manual(values = c("Pre-phase" = "dotted", "Post-phase" = "dashed"),name = "Low investment round") +
  labs(x = "Round",
       y = "Percentage Return",
       color = "Trust Game Phase") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.box = "vertical", # This will stack the legends vertically
        legend.key = element_blank(),
        legend.margin = margin(t = 0.2, b = 0.2, unit = "pt"),
        legend.box.margin = margin(t = 0, b = 0, unit = "pt"),
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  scale_color_manual(values = c("darkblue", "darkgreen"),
                     labels = c("Pre-phase", "Post-phase")) +
  scale_fill_manual(values = c("darkblue", "darkgreen"),
                    guide = "none") + # Hide the fill legend
  guides(color = guide_legend(override.aes = list(fill = NA)))+  # Remove fill for color legend keys
  guides(color = guide_legend(title = "Participant Returns Proportions", override.aes = list(fill = NA)),
         linetype = guide_legend(title = "Low investment round")) +
  facet_wrap(~condition.f)


```

Participants returned higher percentages in the pre phase compared to
the post phase
(`r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f`). This
effect was moderated by Condition
(`r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f`):
contrary to our expectations, participants in the Manipulation condition
decreased their returns from pre to post
(`r papaja::apa_print(prepost_bycond)$full_result$Manipulation_Pre_post`),
while those in the Control condition showed no change (Figure
\@ref(fig:boxPlots)). RS did not moderate this Condition × Phase
interaction.

Higher investments elicited higher percentage returns, indicating
positive reciprocity
(`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$inv_scaled)`).
This relationship was stronger in the Control condition than in the
Manipulation condition
(`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$condition_f_inv_scaled)`).
The effect of investment on returns varied by RS group and Phase
(`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_inv_scaled_high_RS)`),
and a four-way interaction indicated that these patterns further
differed across Conditions
(`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f_inv_scaled_high_RS)`). This model's marginal and conditional $R^2$ were `r {.r <- performance::r2_nakagawa(mod_returns_pct_RS$full_model); sprintf("%.3f and %.3f", .r$R2_marginal, .r$R2_conditional)}`, respectively.

```{r invMod, cache=TRUE}

mod_invs_RS <- mixed( investment ~ phase.f*condition.f*high_RS + (1 + phase.f| playerId), prepost_df, REML= TRUE, method="KR")

summary(mod_invs_RS)
```

```{r}
saveRDS(summary(mod_invs_RS), "data/mod_invs_RS.RDS")
```

```{r violPlotsDef, cache=TRUE}
library(afex)
library(ggplot2)
library(ggsignif)

# Define colors for Phase (different from event study's Condition colors)
phase_cols <- c("pre" = "#E69F00", "post" = "#0072B2")

# Generate the plot data from afex_plot
ret_all_data <- afex_plot(mod_returns_pct_RS, x = "condition.f", trace = "phase.f", dodge = 0.8, error = "within",
                     mapping = c("linetype", "shape", "fill"),
                     factor_levels = list(condition.f = c("Control", "Manipulation")),
                     legend_title = "Phase",
                     emmeans_arg = list(p.adjust.method = "tukey"))

# Extract the data used in afex_plot
plot_data_ret <- ret_all_data$data

# Generate the bar chart for returns
ret_all <- ggplot(plot_data_ret, aes(x = condition.f, y = y, fill = phase.f)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = y - 1.96 * SE, ymax = y + 1.96 * SE),
                position = position_dodge(width = 0.7), width = 0.2, size = 0.5) +
  scale_fill_manual(values = phase_cols, labels = c("Pre-phase", "Post-phase")) +
  theme_bw(base_size = 11) +
  labs(y = "% Returns", x = "Condition", fill = "Trust Game Phase") +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  ) +
  coord_cartesian(ylim = c(0, 0.6)) +
  # Significance bracket for Manipulation pre vs post (moved higher to avoid overlap)
  geom_segment(aes(x = 1.825, xend = 2.175, y = 0.54, yend = 0.54), inherit.aes = FALSE, size = 0.5) +
  geom_segment(aes(x = 1.825, xend = 1.825, y = 0.54, yend = 0.52), inherit.aes = FALSE, size = 0.5) +
  geom_segment(aes(x = 2.175, xend = 2.175, y = 0.54, yend = 0.52), inherit.aes = FALSE, size = 0.5) +
  geom_text(aes(x = 2, y = 0.56, label = "**"), inherit.aes = FALSE, size = 5)


######################################################################################################

# Generate the plot data from afex_plot for investments
inv_all_data <- afex_plot(mod_invs_RS, x = "condition.f", trace = "phase.f", dodge = 0.8, error = "within",
                     mapping = c("linetype", "shape", "fill"),
                     factor_levels = list(condition.f = c("Control", "Manipulation")),
                     legend_title = "Phase")

# Extract the data used in afex_plot
plot_data_inv <- inv_all_data$data

# Generate the bar chart for investments
inv_all <- ggplot(plot_data_inv, aes(x = condition.f, y = y, fill = phase.f)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = y - 1.96 * SE, ymax = y + 1.96 * SE),
                position = position_dodge(width = 0.7), width = 0.2, size = 0.5) +
  scale_fill_manual(values = phase_cols, labels = c("Pre-phase", "Post-phase")) +
  theme_bw(base_size = 11) +
  labs(y = "HMM Investment", x = "Condition", fill = "Trust Game Phase") +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  ) +
  coord_cartesian(ylim = c(0, 12))


```

```{r boxPlots, include=TRUE, fig.pos="H", fig.cap="Marginal means of percentage returns (top) and HMM investments (bottom) by Phase and Condition. Bars show estimated marginal means; error bars represent 95\\% confidence intervals. Participants in the Manipulation condition returned lower proportions post-exposure compared to pre-exposure (** p < .01), while Control participants showed no change. HMM investment did not differ across Phases or Conditions.", fig.align="center", fig.width=5, fig.height=4.5}

library(patchwork)

ret_all / inv_all + plot_layout(guides = 'collect') & theme(legend.position = "bottom", legend.title = element_text(size = 9), legend.text = element_text(size = 8))

```

### Differential responsiveness to investment by RS group

<!-- # USING LMER TO LEVERAGE CAPABILITIES WITH EMTRENDS -->

```{r}
library(lmerTest)
library(emmeans)

# Refitting the model directly
direct_model <- lmer(ret_pct_na ~ phase.f * condition.f * inv_scaled * high_RS + (1 + phase.f | playerId), data = prepost_df, REML = TRUE)



emm_res <- emmeans(direct_model, specs = ~ phase.f * condition.f * high_RS * inv_scaled)
trends <- emtrends(direct_model, specs = ~ phase.f * condition.f * high_RS, var = "inv_scaled")

# Print results
summary(emm_res)
summary(trends)

```

```{r}
library(emmeans)


# First, get the estimated marginal means
emm <- emmeans(mod_returns_pct_RS, ~ inv_scaled | phase.f * condition.f * high_RS)

# Then, get the slopes (effects of investment) for each combination
slopes <- emtrends(mod_returns_pct_RS, ~ phase.f * condition.f * high_RS, var = "inv_scaled")
slopes

# Define the contrasts
corrected_contrasts <- list(
  "ManipLowRS" = c(0, 0, -1, 1, 0, 0, 0, 0),
  "ManipHighRS" = c(0, 0, 0, 0, 0, 0, -1, 1),
  "HighVsLowRS" = c(0, 0, 1, -1, 0, 0, -1, 1)
)

# Test the contrasts with multiple comparison correction
results_4w <- contrast(slopes, method = corrected_contrasts, adjust = "sidak")

# View the results
summary(results_4w)

# Save slopes and contrasts for supplement
saveRDS(list(slopes = summary(slopes), contrasts = summary(results_4w)), "data/mod_slopes_4w.RDS")

```

To examine this four-way interaction, we conducted a contrast analysis
of how the effect of investment on returns changed from pre- to
post-exposure for different RS groups in both conditions (Figure \@ref(fig:fourwayInteractionPlot)). Starting
with the Manipulation condition, for participants with low RS, the
effect of investment on returns increased significantly from pre- to
post-phase,
`r papaja::apa_print(results_4w)$full_result$ManipLowRS`.
This suggests that after the manipulation, low RS participants became
more responsive to their co-player's investments, returning
proportionally more as investments increased. In contrast, for
participants with high RS, the effect of investment on returns decreased
significantly from pre- to post-exposure,
`r papaja::apa_print(results_4w)$full_result$ManipHighRS`.
This indicates that high RS participants became less responsive to their
co-player's investments after the manipulation, with smaller increases
in returns as investments increased. The difference in these pre-post
changes between high and low RS groups was significant,
`r papaja::apa_print(results_4w)$full_result$HighVsLowRS`.
This result suggests that the manipulation had significantly different
effects on how low and high RS participants responded to their
co-player's investments.

In the Control condition, we observed no significant changes in how
participants responded to their co-player's investments between the pre
and post phases, regardless of their RS level.

```{r}
# Define the contrasts for both conditions
contrasts_ctrl <- list(
  # Control condition contrasts
  "Diff in inv effect (Post - Pre, Control, Low RS)" = c(-1, 1, 0, 0, 0, 0, 0, 0),
  "Diff in inv effect (Post - Pre, Control, High RS)" = c(0, 0, 0, 0, -1, 1, 0, 0),
  "Diff in (Post - Pre) between High and Low RS in Control" = c(1, -1, 0, 0, -1, 1, 0, 0)
)

# Test the contrasts with multiple comparison correction
results_ctrl <- contrast(slopes, method = contrasts_ctrl, adjust = "sidak")

# View the results
print(summary(results_ctrl))
```

```{r fourwayInteractionPlot, include=TRUE, fig.pos="H", fig.width=5.5, fig.height=3.15, fig.cap ="Marginal effect of investment on percentage returns by Phase, Condition, and RS group. Bars show estimated slopes (change in returns per SD increase in investment) from the mixed model; error bars represent 95\\% confidence intervals. In the Manipulation condition, Low RS participants became more responsive to investments post-exposure (* p < .05), while High RS participants became less responsive (* p < .05). No significant changes were observed in the Control condition."}

library(ggplot2)

# Convert emtrends output to data frame for plotting
slopes_df <- as.data.frame(summary(slopes))

# Rename columns for clarity
slopes_df <- slopes_df %>%
  mutate(
    Phase = factor(phase.f, levels = c("pre", "post"), labels = c("Pre", "Post")),
    Condition = factor(condition.f, levels = c("Control", "Manipulation")),
    RS_Group = factor(high_RS, levels = c("low RS", "high RS"), labels = c("Low RS", "High RS"))
  )

# Define colors matching Figure 4 (Phase colors)
phase_cols <- c("Pre" = "#E69F00", "Post" = "#0072B2")

# Create the bar chart
ggplot(slopes_df, aes(x = Condition, y = inv_scaled.trend, fill = Phase)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, color = "black", size = 0.3) +
  geom_errorbar(
    aes(ymin = inv_scaled.trend - 1.96 * SE, ymax = inv_scaled.trend + 1.96 * SE),
    position = position_dodge(width = 0.7),
    width = 0.2,
    size = 0.5
  ) +
  facet_wrap(~ RS_Group, ncol = 2) +
  scale_fill_manual(values = phase_cols, labels = c("Pre-phase", "Post-phase")) +
  labs(
    x = "Condition",
    y = "Effect of Investment on Returns\n(Slope ± 95% CI)",
    fill = "Phase"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  ) +
  # Add significance bracket for Low RS panel (Manipulation: Pre vs Post)
  geom_segment(
    data = data.frame(RS_Group = factor("Low RS", levels = c("Low RS", "High RS"))),
    aes(x = 1.825, xend = 2.175, y = 0.068, yend = 0.068),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_segment(
    data = data.frame(RS_Group = factor("Low RS", levels = c("Low RS", "High RS"))),
    aes(x = 1.825, xend = 1.825, y = 0.068, yend = 0.065),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_segment(
    data = data.frame(RS_Group = factor("Low RS", levels = c("Low RS", "High RS"))),
    aes(x = 2.175, xend = 2.175, y = 0.068, yend = 0.065),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_text(
    data = data.frame(RS_Group = factor("Low RS", levels = c("Low RS", "High RS"))),
    aes(x = 2, y = 0.072, label = "*"),
    inherit.aes = FALSE, size = 6
  ) +
  # Add significance bracket for High RS panel (Manipulation: Pre vs Post)
  geom_segment(
    data = data.frame(RS_Group = factor("High RS", levels = c("Low RS", "High RS"))),
    aes(x = 1.825, xend = 2.175, y = 0.058, yend = 0.058),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_segment(
    data = data.frame(RS_Group = factor("High RS", levels = c("Low RS", "High RS"))),
    aes(x = 1.825, xend = 1.825, y = 0.058, yend = 0.055),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_segment(
    data = data.frame(RS_Group = factor("High RS", levels = c("Low RS", "High RS"))),
    aes(x = 2.175, xend = 2.175, y = 0.058, yend = 0.055),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_text(
    data = data.frame(RS_Group = factor("High RS", levels = c("Low RS", "High RS"))),
    aes(x = 2, y = 0.062, label = "*"),
    inherit.aes = FALSE, size = 6
  )

```

### Sensitivity to suspicion of AI agent

```{r suspicionSensitivity, include=FALSE, cache=TRUE}
# -----------------------------------------------------------------------------
# Sensitivity analysis added in response to Reviewer 1, comment 5
# (Collabra R&R, May 2026): re-fit the headline Phase x Condition x Investment
# x RS returns model on subsets defined by the post-experiment Turing question.
# -----------------------------------------------------------------------------

# Join Turing.choice from d_finished onto the existing prepost_df.
turing_lookup <- d_finished %>%
  dplyr::distinct(playerId, Turing.choice) %>%
  dplyr::filter(!is.na(Turing.choice))

prepost_susp_df <- prepost_df %>%
  dplyr::inner_join(turing_lookup, by = "playerId") %>%
  dplyr::mutate(Turing.choice = factor(Turing.choice,
                                       levels = c("Human", "Machine", "Not Sure")))

# Participant-level counts and percentages
turing_counts <- prepost_susp_df %>%
  dplyr::distinct(playerId, .keep_all = TRUE) %>%
  dplyr::count(Turing.choice)
n_human   <- turing_counts$n[turing_counts$Turing.choice == "Human"]
n_machine <- turing_counts$n[turing_counts$Turing.choice == "Machine"]
n_notsure <- turing_counts$n[turing_counts$Turing.choice == "Not Sure"]
n_total   <- sum(turing_counts$n)
pct_human    <- round(n_human   / n_total * 100)
pct_machine  <- round(n_machine / n_total * 100)
pct_notsure  <- round(n_notsure / n_total * 100)
# "Detected" = answered Machine; "Fooled" = answered Human or Not Sure (the agent's
# cover held: these participants did not correctly identify a machine).
pct_detected <- pct_machine
pct_fooled   <- round((n_human + n_notsure) / n_total * 100)

# Pooled sensitivity model: detection status (answered Machine vs Human/Not
# Sure) enters as a between-subjects factor in a single returns model. This
# yields both the Manipulation pre-to-post contrast within each detection
# group (via emmeans, with 95% CIs) and a formal test of whether the
# manipulation effect depends on detection (the Phase x Condition x Detection
# interaction). Investment and RS terms are omitted: the question here is
# only whether the primary Phase x Condition effect depends on detection; the
# RS four-way is reported for the full sample and is not powered for
# subgroups.
prepost_susp_df <- prepost_susp_df %>%
  dplyr::mutate(detection = factor(
    ifelse(Turing.choice == "Machine", "detected", "not undeceived"),
    levels = c("detected", "not undeceived")))

n_detected <- dplyr::n_distinct(prepost_susp_df$playerId[prepost_susp_df$detection == "detected"])
n_fooled   <- dplyr::n_distinct(prepost_susp_df$playerId[prepost_susp_df$detection == "not undeceived"])

mod_susp <- mixed(
  ret_pct_na ~ phase.f * condition.f * detection + (1 + phase.f | playerId),
  prepost_susp_df, REML = TRUE, method = "KR")

# Manipulation pre-to-post contrast within each detection group, with CIs.
emm_susp <- emmeans(mod_susp, ~ phase.f | condition.f * detection,
                    pbkrtest.limit = 9000, lmerTest.limit = 9000)
susp_pairs <- summary(pairs(emm_susp), infer = c(TRUE, TRUE))
susp_det <- susp_pairs[susp_pairs$condition.f == "Manipulation" &
                         susp_pairs$detection == "detected", ]
susp_fooled <- susp_pairs[susp_pairs$condition.f == "Manipulation" &
                            susp_pairs$detection == "not undeceived", ]

# Phase x Condition x Detection interaction from the Kenward-Roger table.
susp_3way <- mod_susp$anova_table["phase.f:condition.f:detection", ]

# Cache for faster re-renders; susp_summary feeds the supplement's
# coefficient table.
saveRDS(list(
  turing_counts = turing_counts,
  susp_pairs    = susp_pairs,
  susp_3way     = susp_3way,
  susp_summary  = summary(mod_susp),
  n_detected    = n_detected,
  n_fooled      = n_fooled
), "data/mod_suspicion_sensitivity.RDS")
```

Because participants were not informed that they faced computerized
investors, at the end of the experiment we asked whether they believed
they had faced humans, machines, or were unsure (the "Turing
question"). Of the `r n_total` participants, `r pct_human`% believed
they faced humans, `r pct_notsure`% were unsure, and `r pct_machine`%
correctly identified a machine. Because perceived partner identity can
itself shape cooperation strategies [@liu_social_2025], a natural
concern is whether the
primary, powered result, the post-exposure reduction in returns, is
confined to participants who saw through the agent. We therefore added
detection status, whether the participant detected the agent (the
`r pct_detected`% who answered "machine") or did not (the remaining
`r pct_fooled`%, who answered "human" or were unsure and so were not
undeceived by the cover story), as a between-subjects factor to a
returns model with Phase and Condition. The Manipulation pre-to-post
decrease in returns was of similar magnitude in both groups (detected,
$n$ = `r n_detected`: $\Delta M$ =
`r sprintf("%.3f", susp_det$estimate)`, 95% CI
[`r sprintf("%.3f", susp_det$lower.CL)`,
`r sprintf("%.3f", susp_det$upper.CL)`], $p$ =
`r papaja::printp(susp_det$p.value)`; not undeceived, $n$ =
`r n_fooled`: $\Delta M$ = `r sprintf("%.3f", susp_fooled$estimate)`,
95% CI [`r sprintf("%.3f", susp_fooled$lower.CL)`,
`r sprintf("%.3f", susp_fooled$upper.CL)`], $p$ =
`r papaja::printp(susp_fooled$p.value)`), and there was no evidence
that the manipulation effect depended on detection (Phase $\times$
Condition $\times$ Detection:
$F$(`r round(susp_3way[["num Df"]])`,
`r sprintf("%.2f", susp_3way[["den Df"]])`) =
`r sprintf("%.2f", susp_3way[["F"]])`, $p$ =
`r papaja::printp(susp_3way[["Pr(>F)"]])`), although this interaction
test is not separately powered. The reduction therefore does not
depend on whether participants believed their partner was human. The
higher-order interaction with rejection sensitivity is reported for the
full sample above; we did not subdivide it by suspicion status, as a
four-way interaction is not adequately powered within these subgroups.

### Exploitation diagnostic

```{r exploitDiagnostic, include=FALSE, cache=TRUE}
# -----------------------------------------------------------------------------
# Create exploitation variable: return < 1/3 means investor loses money
# -----------------------------------------------------------------------------
exploit_df <- prepost_df %>%
  filter(!is.na(ret_pct_na)) %>%
  mutate(
    exploit = as.numeric(ret_pct_na < 1/3),
    trustee_payoff = (3 * investment) - returns,
    investor_payoff = returns - investment
  )

# -----------------------------------------------------------------------------
# Model: Mixed logistic regression for exploitation probability
# -----------------------------------------------------------------------------
mod_exploit <- glmer(
  exploit ~ phase.f * condition.f + inv_scaled + high_RS + (1 | playerId),
  data = exploit_df,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa")
)
saveRDS(summary(mod_exploit), "data/mod_exploit.RDS")

# Get estimated marginal means for Phase × Condition
emm_exploit <- emmeans(mod_exploit, ~ phase.f * condition.f, type = "response")
exploit_summary <- as.data.frame(summary(emm_exploit))

# Test the Phase × Condition interaction
exploit_interaction <- contrast(emm_exploit, interaction = "pairwise")
exploit_interaction_summary <- summary(exploit_interaction)

# Pre vs Post within each condition
exploit_bycond <- pairs(emmeans(mod_exploit, "phase.f", by = "condition.f", type = "response"))

# -----------------------------------------------------------------------------
# Payoff analysis
# -----------------------------------------------------------------------------
# Manipulation condition payoffs
manip_pre_payoff <- exploit_df %>% filter(condition.f == "Manipulation", phase.f == "pre") %>%
  group_by(playerId) %>% summarise(payoff = mean(trustee_payoff), .groups = "drop") %>% pull(payoff)
manip_post_payoff <- exploit_df %>% filter(condition.f == "Manipulation", phase.f == "post") %>%
  group_by(playerId) %>% summarise(payoff = mean(trustee_payoff), .groups = "drop") %>% pull(payoff)
t_manip_payoff <- t.test(manip_pre_payoff, manip_post_payoff, paired = TRUE)

# -----------------------------------------------------------------------------
# Between-condition trustee payoff comparison in the post phase.
# Added in response to Reviewer 1, comment 7 (Collabra R&R, May 2026):
# direct between-condition test that complements the within-Manipulation
# pre/post test above and the exploitation-probability Phase x Condition
# interaction further up. If Manipulation participants had been
# strategically exploiting the post-exposure agent, their per-round
# trustee payoff (= 3*investment - returns, i.e., what they keep) should
# exceed Control's.
# -----------------------------------------------------------------------------
ctrl_post_payoff <- exploit_df %>% filter(condition.f == "Control", phase.f == "post") %>%
  group_by(playerId) %>% summarise(payoff = mean(trustee_payoff), .groups = "drop") %>% pull(payoff)
t_post_between  <- t.test(manip_post_payoff, ctrl_post_payoff, paired = FALSE)
w_post_between  <- wilcox.test(manip_post_payoff, ctrl_post_payoff, paired = FALSE)
mean_manip_post <- mean(manip_post_payoff)
mean_ctrl_post  <- mean(ctrl_post_payoff)
sd_manip_post   <- sd(manip_post_payoff)
sd_ctrl_post    <- sd(ctrl_post_payoff)
n_manip_post    <- length(manip_post_payoff)
n_ctrl_post     <- length(ctrl_post_payoff)

# Descriptive exploitation rates
exploit_rates <- exploit_df %>%
  group_by(phase.f, condition.f) %>%
  summarise(exploit_rate = mean(exploit) * 100, .groups = "drop")
```

A mixed logistic regression predicting exploitation probability (returns below one-third of the tripled investment) revealed that exploitation increased from pre to post in both conditions (main effect of Phase, $z$ = `r round(summary(mod_exploit)$coefficients["phase.fpost", "z value"], 2)`, $p$ = `r printp(summary(mod_exploit)$coefficients["phase.fpost", "Pr(>|z|)"])`). The Phase × Condition interaction was not significant ($z$ = `r sprintf("%.2f", exploit_interaction_summary$z.ratio)`, $p$ = `r printp(exploit_interaction_summary$p.value)`). The model's marginal and conditional $R^2$ were `r {.r <- performance::r2_nakagawa(mod_exploit, approximation = "lognormal"); sprintf("%.3f and %.3f", .r$R2_marginal, .r$R2_conditional)}`, respectively. The Manipulation condition showed lower exploitation rates than Control at both time points (pre: `r round(exploit_rates$exploit_rate[exploit_rates$phase.f == "pre" & exploit_rates$condition.f == "Manipulation"], 1)`% vs. `r round(exploit_rates$exploit_rate[exploit_rates$phase.f == "pre" & exploit_rates$condition.f == "Control"], 1)`%; post: `r round(exploit_rates$exploit_rate[exploit_rates$phase.f == "post" & exploit_rates$condition.f == "Manipulation"], 1)`% vs. `r round(exploit_rates$exploit_rate[exploit_rates$phase.f == "post" & exploit_rates$condition.f == "Control"], 1)`%). Trustee payoffs in the Manipulation condition did not significantly change from pre ($M$ = `r round(mean(manip_pre_payoff), 1)`) to post ($M$ = `r round(mean(manip_post_payoff), 1)`; $t$(`r t_manip_payoff$parameter`) = `r round(t_manip_payoff$statistic, 2)`, $p$ = `r printp(t_manip_payoff$p.value)`). A direct between-condition comparison in the post phase further supports the absence of exploitation: per-round trustee payoffs in Manipulation ($M$ = `r round(mean_manip_post, 2)`, $SD$ = `r round(sd_manip_post, 2)`, $n$ = `r n_manip_post`) did not differ from those in Control ($M$ = `r round(mean_ctrl_post, 2)`, $SD$ = `r round(sd_ctrl_post, 2)`, $n$ = `r n_ctrl_post`; Welch $t$(`r round(t_post_between$parameter, 1)`) = `r round(t_post_between$statistic, 2)`, $p$ = `r printp(t_post_between$p.value)`; Mann-Whitney $W$ = `r w_post_between$statistic`, $p$ = `r printp(w_post_between$p.value)`). If exposure had led participants to exploit the post-exposure agent, Manipulation trustees should have kept more per round than Control trustees; they did not.

### Returns prior to pre-programmed low investment trials

```{r preDefectionData, include=FALSE}
# Prepare data for rounds prior to pre-programmed low investment
# Pre-phase: rounds 1-11 (low investment at round 12)
# Post-phase: rounds 1-12 (low investment at round 13)
predef_df <- avg_ret_df %>%
  filter(
    (gameNum.f == "pre" & roundNum < 12) |
    (gameNum.f == "post" & roundNum < 13)
  )
```

```{r preDefectionModel, include=FALSE, cache=TRUE}
# Same model as main analysis but restricted to rounds prior to low investment
mod_predef <- mixed(ret_pct_na ~ phase.f * condition.f * inv_scaled * high_RS + (1 + phase.f | playerId),
                    data = predef_df, REML = TRUE, method = "KR")
saveRDS(summary(mod_predef), "data/mod_predef.RDS")
```

```{r preDefectionContrasts, include=FALSE, cache=TRUE}
# Pre vs Post within each Condition
prepost_bycond_predef <- pairs(emmeans(mod_predef, "phase.f", by = "condition.f", pbkrtest.limit = 5000))
```

To distinguish between contrast effects and betrayal aversion as explanations for reduced cooperation in the Manipulation condition, we examined returns in the rounds preceding the pre-programmed low investment. If contrast effects were operating, participants in the Manipulation condition should already show reduced returns before experiencing any low investment in the post-exposure phase. Conversely, if betrayal aversion were the primary mechanism, group differences should only emerge after the low investment.

The Phase $\times$ Condition interaction was significant in rounds prior to the low investment (`r papaja::apa_print(mod_predef)$full_result$phase_f_condition_f`). Participants in the Manipulation condition significantly decreased their returns from pre- to post-exposure phase even before encountering the low investment (`r papaja::apa_print(prepost_bycond_predef)$full_result$Manipulation_Pre_post`), whereas those in the Control condition showed no change (`r papaja::apa_print(prepost_bycond_predef)$full_result$Control_Pre_post`). The four-way interaction also remained significant (`r papaja::apa_print(mod_predef)$full_result$phase_f_condition_f_inv_scaled_high_RS`), suggesting that the differential responsiveness to investments observed in the full analysis was likewise present before the low investment occurred. This model's marginal and conditional $R^2$ were `r {.r <- performance::r2_nakagawa(mod_predef$full_model); sprintf("%.3f and %.3f", .r$R2_marginal, .r$R2_conditional)}`, respectively.

<!-- Focusing on post defection only:  -->

### Reaction to pre-programmed low investment: event study analysis

To understand how participants reacted to and recovered from the pre-programmed low investment, we examined the Drop and Recovery (defined above) in the event study centered on the low investment round (Figure \@ref(fig:eventStudyPlot)); negative Drop values indicate reduced returns (punishment of low trust), and positive Recovery values indicate restored cooperation.



```{r eventStudyData, include=FALSE}
# Define defection rounds for each phase
def_rounds <- data.frame(
  gameNum.f = c("pre", "post"),
  def_round = c(12, 13)
)

# Prepare data: Calculate relative round (t=0 is defection)
event_data <- avg_ret_df %>%
  filter(gameNum.f %in% c("pre", "post")) %>%
  mutate(gameNum.f = as.character(gameNum.f)) %>% # Ensure matching types for join
  left_join(def_rounds, by = "gameNum.f") %>%
  mutate(
    rel_round = roundNum - def_round,
    # Create a factor for plotting order with nice labels
    time_point = factor(rel_round, levels = -2:2, labels = c("t_minus_2", "t_minus_1", "Defection", "t_plus_1", "t_plus_2"))
  ) %>%
  filter(rel_round >= -2 & rel_round <= 2)
```

```{r eventStudyModel, include=FALSE, cache=TRUE}
# Fit mixed model with RS moderation
# We use a full interaction model to test if the trajectory (time_point) differs by Condition, Phase, and RS
mod_event <- mixed(ret_pct_na ~ phase.f * condition.f * time_point * high_RS + (1 | playerId), 
                   data = event_data, 
                   REML = TRUE, 
                   method = "KR")

# Setup contrasts for "Drop" (t-1 to Defection) and "Recovery" (Defection to t+1)
emm_event <- emmeans(mod_event, ~ time_point | phase.f * condition.f * high_RS)

custom_contrasts <- list(
  "Drop"     = c(0, -1, 1, 0, 0),    # Defection - t_minus_1 (negative when returns decrease)
  "Recovery" = c(0, 0, -1, 1, 0)     # t_plus_1 - Defection (positive when returns recover)
)

# Calculate specific contrasts
event_contrasts <- contrast(emm_event, custom_contrasts)

# --- NEW: Calculate contrasts irrespective of RS (Condition x Phase) ---
# We average over RS levels to get the main effect of Condition in each Phase
emm_event_noRS <- emmeans(mod_event, ~ time_point | phase.f * condition.f)
event_contrasts_noRS <- contrast(emm_event_noRS, custom_contrasts)
cond_diffs_post_noRS <- contrast(event_contrasts_noRS, interaction = c("identity", "pairwise", "identity"), by = "phase.f")

# Test differences between conditions in the Post phase (RS specific)
cond_diffs_post <- contrast(event_contrasts, interaction = c("identity", "pairwise", "identity"), by = "phase.f")

# R1.4 / R2.5 (Collabra R&R, May 2026): level-difference contrast (Control -
# Manipulation) at t+2 in the post-exposure, high-RS cell. Tests whether the
# between-condition gap has closed by t+2, the longest post-breach window the
# design permits (breach at round 13 of 15).
emm_event_level <- emmeans(mod_event, ~ condition.f | phase.f * time_point * high_RS,
                            pbkrtest.limit = 3000, lmerTest.limit = 3000)
level_diffs_df <- as.data.frame(pairs(emm_event_level))
hi_post_t2_gap <- level_diffs_df %>%
  dplyr::filter(phase.f == "post", high_RS == "high RS", time_point == "t_plus_2")

# Extract specific p-values and estimates for reporting
# We need to grab these values to use in inline R
# Helper function or just manual extraction
get_p_val <- function(contrasts_obj, row_idx) {
  s <- summary(contrasts_obj)
  p <- s$p.value[row_idx]
  if(p < .001) return("< .001")
  return(paste("=", round(p, 3)))
}
saveRDS(summary(mod_event), "data/mod_event.RDS")
```

```{r eventStudyPlot, echo=FALSE, include=TRUE, fig.pos="H", fig.width=5.5, fig.height=3.15, fig.cap="Drop and Recovery responses to pre-programmed low investment in the post-exposure phase (pre-exposure phase not shown as no between-condition differences were observed). Drop = change in returns from low investment round minus t-1 (negative values indicate reduced returns); Recovery = change from t+1 minus low investment round (positive values indicate restored returns). Bars show estimated marginal means from the mixed model; error bars represent 95\\% confidence intervals. The bracket shows the significant between-condition difference in Recovery for High RS participants. ** p < .01."}

# Extract contrast estimates from the model for Post phase only
contrast_summary <- summary(event_contrasts) %>%
  as.data.frame() %>%
  filter(phase.f == "post") %>%
  mutate(
    # Clean up names for plotting
    Contrast = factor(contrast, levels = c("Drop", "Recovery")),
    Condition = factor(condition.f, levels = c("Control", "Manipulation")),
    RS_Group = factor(high_RS, levels = c("low RS", "high RS"))
  )

# Extract between-condition significance tests for Post phase
cond_diffs_summary <- summary(cond_diffs_post) %>%
  as.data.frame() %>%
  filter(phase.f == "post") %>%
  mutate(
    sig_label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "n.s."
    )
  )

# Define colors
cols <- c("Control" = "#F8766D", "Manipulation" = "#00BFC4")

# Create the bar chart
p <- ggplot(contrast_summary, aes(x = Contrast, y = estimate, fill = Condition)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, color = "black", size = 0.3) +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * SE, ymax = estimate + 1.96 * SE),
    position = position_dodge(width = 0.7),
    width = 0.2,
    size = 0.5
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", size = 0.5) +
  facet_wrap(~ RS_Group, ncol = 2) +
  scale_fill_manual(values = cols) +
  labs(
    x = "Behavioral Response",
    y = "Change in Percentage Return",
    fill = "Condition"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  ) +
  # Add significance bracket for Recovery in High RS panel
  # Get the y-position for the bracket (above the higher bar)
  geom_segment(
    data = data.frame(RS_Group = factor("high RS", levels = c("low RS", "high RS"))),
    aes(x = 1.825, xend = 2.175, y = 0.18, yend = 0.18),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_segment(
    data = data.frame(RS_Group = factor("high RS", levels = c("low RS", "high RS"))),
    aes(x = 1.825, xend = 1.825, y = 0.18, yend = 0.17),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_segment(
    data = data.frame(RS_Group = factor("high RS", levels = c("low RS", "high RS"))),
    aes(x = 2.175, xend = 2.175, y = 0.18, yend = 0.17),
    inherit.aes = FALSE, size = 0.5
  ) +
  geom_text(
    data = data.frame(RS_Group = factor("high RS", levels = c("low RS", "high RS"))),
    aes(x = 2, y = 0.20, label = "**"),
    inherit.aes = FALSE, size = 5
  )

print(p)
```


In the pre-exposure phase, there were no significant differences between conditions in either the Drop (`r papaja::apa_print(cond_diffs_post_noRS)$full_result$Drop_Control_Manipulation_Pre`) or Recovery (`r papaja::apa_print(cond_diffs_post_noRS)$full_result$Recovery_Control_Manipulation_Pre`), confirming that both groups started with equivalent behavioral patterns. In the post-exposure phase, a divergence emerged. The Control group showed a significantly larger Drop than the Manipulation group (`r papaja::apa_print(cond_diffs_post_noRS)$full_result$Drop_Control_Manipulation_Post`), indicating that participants exposed to the forgiving agent showed a blunted immediate reaction to the low investment and did not reduce their returns as sharply. The subsequent Recovery did not differ significantly between conditions overall (`r papaja::apa_print(cond_diffs_post_noRS)$full_result$Recovery_Control_Manipulation_Post`). The event-study model's marginal and conditional $R^2$ were `r {.r <- performance::r2_nakagawa(mod_event$full_model); sprintf("%.3f and %.3f", .r$R2_marginal, .r$R2_conditional)}`, respectively.

When examining moderation by RS, the pattern appeared to be driven primarily by high RS participants. Low RS participants showed no significant difference in Recovery between conditions (`r papaja::apa_print(cond_diffs_post)$full_result$Recovery_Control_Manipulation_Low_RS_Post`), with both groups displaying modest, similar recovery patterns after the low investment. The larger Drop observed in the Control condition for this group may partly reflect their elevated cooperation level at $t-1$.

High RS participants showed a different pattern. In the Control condition, they demonstrated trust repair by significantly increasing their returns after the low investment (`r papaja::apa_print(event_contrasts)$full_result$Post_Control_High_RS_Recovery`). However, those in the Manipulation condition did not show immediate recovery, with no significant increase in returns at $t+1$ (`r papaja::apa_print(event_contrasts)$full_result$Post_Manipulation_High_RS_Recovery`). The difference in Recovery between conditions was significant (`r papaja::apa_print(cond_diffs_post)$full_result$Recovery_Control_Manipulation_High_RS_Post`). By $t+2$, however, the Control-Manipulation gap in high RS participants had effectively closed ($\Delta$ = `r sprintf("%.3f", hi_post_t2_gap$estimate)`, $t$(`r round(hi_post_t2_gap$df)`) = `r sprintf("%.2f", hi_post_t2_gap$t.ratio)`, $p$ = `r papaja::printp(hi_post_t2_gap$p.value)`), suggesting that the recovery is delayed rather than absent. The late placement of the programmed breach (round 13 of 15) limits how far we can track this trajectory, and a design with an earlier breach and longer post-breach window would be needed to characterize the full course of trust repair in high RS individuals.

In summary, exposure to the forgiving agent appeared to dampen reciprocal responsiveness and to delay the re-establishment of cooperation following a temporary withdrawal of trust; this delay was transient. The pattern was more pronounced among high RS individuals: whereas high RS participants in the Control condition reduced returns sharply when trust was withdrawn and increased them immediately afterward, those exposed to the forgiving agent showed a muted reaction to the low investment and a one-round delay before reinstating higher returns.

<!-- Old Event Study line plot - replaced by bar chart above
```{r eventStudy, eval=FALSE, include=FALSE, fig.cap="Event study of participant returns centered on the defection round (t=0). Comparison of Control and Manipulation conditions across Pre and Post phases."}
# Define defection rounds for each phase
def_rounds <- data.frame(
  gameNum.f = c("pre", "post"),
  def_round = c(12, 13)
)

# Prepare data: Calculate relative round (t=0 is defection)
event_data <- avg_ret_df %>%
  filter(gameNum.f %in% c("pre", "post")) %>%
  mutate(gameNum.f = as.character(gameNum.f)) %>% # Ensure matching types for join
  left_join(def_rounds, by = "gameNum.f") %>%
  mutate(
    rel_round = roundNum - def_round,
    # Create a factor for plotting order with nice labels
    time_point = factor(rel_round, levels = -2:2, labels = c("t-2", "t-1", "Betrayal", "t+1", "t+2"))
  ) %>%
  filter(rel_round >= -2 & rel_round <= 2) # Filter for the window of interest

# Plot the event study
ggplot(event_data, aes(x = time_point, y = ret_pct_na, color = condition.f, group = condition.f)) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  facet_wrap(~phase.f) +
  labs(
    title = "Reaction to Defection (Event Study)",
    x = "Trial relative to Defection",
    y = "Percentage Return",
    color = "Condition"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```
-->

<!-- 
### Post Defection Trials

```{r postDefMod, cache=TRUE}
post_def_data <- avg_ret_df %>% filter(roundType=="trust",!is.na(gameNum.f),(roundNum >= 12 & gameNum.f =="pre")| (roundNum >= 13 & gameNum.f =="post"))

mod_retPostDef_RS <- mixed( ret_pct_na ~ phase.f*condition.f*inv_scaled + (1 + phase.f| playerId), post_def_data , REML= TRUE, method="KR")

summary(mod_retPostDef_RS)
```

```{r PairsPostDef}
phasePairs <- pairs(emmeans::emmeans(mod_retPostDef_RS, c("phase.f"), pbkrtest.limit = 1449))
phasePairs
```

<!-- Did participants learn to be more forgiving and cooperative after -->
<!-- witnessing the pre-programmed defection by the HMM investor? To explore -->
<!-- this question, we restrict the analysis to the trials following the -->
<!-- pre-programmed defection by the HMM agent in both the "pre" (trials 12 -->
<!-- to 15) and the "post" phases (trials 13 to 15). We fit the same mixed -->
<!-- effects model as for all the trials with the exception of the RS -->
<!-- variable. This is because RS did not show main effects in the previous -->
<!-- model, and also due to the necessity of running a simpler model to -->
<!-- accommodate the low number of trials. We find a significant main effect -->
<!-- of Phase `r papaja::apa_print(mod_retPostDef_RS)$full_result$phase_f` -->
<!-- with returns lower in the second game post defection trials compared to -->
<!-- the first. We also find a main effect of Investment -->
<!-- `r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_retPostDef_RS)$full_result$inv_scaled)` -->
<!-- where participants continued to return higher proportions when receiving -->
<!-- higher investments. Finally, we still find an Investment by Condition -->
<!-- interaction -->
<!-- `r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$condition_f_inv_scaled)` -->
<!-- showing a lower effects of investment on the Manipulation condition -->
<!-- compared to the Control condition in post-defection trials. However, the -->
<!-- absence of a Condition by Phase effect indicates there was no difference -->
<!-- between conditions on participants' reaction to a one-off low investment -->
<!-- by the co-player. -->
<!-- -->

<!-- Checking whether Investor behavior has changed dramatically between pre and post: -->

### HMM investor in pre and post phases

<!-- Linear mixed effects model for investments -->

Was the HMM's strategy similar between pre and post phases in the
control condition? Was participants' behavior post exposure
differentiated enough to induce a different reaction from the HMM? To
answer these questions, we test for differences in the HMM agent's
investment by Phase, Condition and RS using a linear mixed-effects model
as described in the methods section. As seen in Figure
\@ref(fig:boxPlots), we find no main or interaction effects, indicating
the HMM's behavior was on aggregate similar across levels of Phase,
Condition and RS. This consistency in the investor's behavior is a
desirable feature of the HMM agent when the participants' behavior is
largely similar between phases. More importantly, it indicates that the
lower returns of participants in the post phase of the manipulation
condition were not differentiated enough to make the HMM react by
transitioning to lower latent trust states. It is also noteworthy that
several participants engaged with the agent as though it were a human
partner: when asked to justify their Turing-question responses, many
projected human traits such as "spitefulness" or "greed" onto the
co-player's behavior (the full breakdown of human, machine, and unsure
responses is reported in "Sensitivity to suspicion of AI agent").

### Exposure phase trials

```{r}
ggplot(avg_ret_df %>% filter(gameNum.f %in% c("expo1","expo2","expo3")), aes(x=as.factor(roundNum), y=ret_pct_na, group=gameNum.f, color = gameNum.f, fill=gameNum.f)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(ymin=..ymin.., ymax=..ymax..),
               alpha = 0.3, linetype = 0) +
  # geom_vline(data = vline_data, aes(xintercept = xintercept, linetype = Defection_round), color = "black", linewidth = 0.6) +
  # scale_linetype_manual(values = c("Pre-phase" = "dotted", "Post-phase" = "dashed"),name = "Defection round") +
  labs(x = "Round",
       y = "Percentage Return",
       color = "Trust Game Phase") +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(fill = NA)))+  # Remove fill for color legend keys
  guides(color = guide_legend(title = "Participant Returns Proportions", override.aes = list(fill = NA)),
         linetype = guide_legend(title = "Low investment round")) +
  facet_wrap(~condition.f)
```

```{r}

expo_df <- avg_ret_df %>% filter(gameNum.f %in% c("expo1","expo2","expo3"))

mod_returns_expo <- mixed( ret_pct_na ~ condition.f*inv_scaled*high_RS + (1 | playerId), expo_df, REML= TRUE, method="KR")
saveRDS(summary(mod_returns_expo), "data/mod_returns_expo.RDS")
mod_inv_summary <- summary(mod_returns_expo)
mod_inv_summary

em_ret_cond <- emmeans::emmeans(mod_returns_expo , c("condition.f"))
pairs(em_ret_cond)

# Calculate the slopes (coefficients) of inv_scaled for each condition
inv_slopes <- emtrends(mod_returns_expo, ~ condition.f, var = "inv_scaled")

# To see pairwise comparisons of the slopes
print(inv_slopes)
pairs(inv_slopes)

##### Three way interaction
inv_slopes3 <- emtrends(mod_returns_expo, ~ condition.f * high_RS, var = "inv_scaled")
summary(inv_slopes3)

pairwise_comparisons3 <- pairs(inv_slopes3, by = "high_RS")
summary(pairwise_comparisons3)


######### HMM investment in EXPO PHASE
mod_invs_expo <- mixed(investment ~ condition.f*high_RS + (1 | playerId), expo_df, REML= TRUE, method="KR")
saveRDS(summary(mod_invs_expo), "data/mod_invs_expo.RDS")
summary(mod_invs_expo)

emcheck_inv <- pairs(emmeans::emmeans(mod_invs_expo, c("condition.f"), pbkrtest.limit = 6300))
emcheck_inv
```

```{r exposureAbsoluteReturns, include=FALSE, cache=TRUE}
# -----------------------------------------------------------------------------
# Absolute returns during the exposure phase.
# Added in response to Reviewer 1, comment 3 (Collabra R&R, May 2026):
# percentage returns were similar across conditions but investments were
# higher in Manipulation; reviewer asks whether the implied absolute
# returns differed. Marginal mixed model (no inv covariate) gives the
# directly interpretable per-condition mean of absolute returns.
# -----------------------------------------------------------------------------

mod_returns_abs_expo <- afex::mixed(
  returns ~ condition.f * high_RS + (1 | playerId),
  data = expo_df, REML = TRUE, method = "KR")
saveRDS(summary(mod_returns_abs_expo), "data/mod_returns_abs_expo.RDS")

anova_abs_expo <- anova(mod_returns_abs_expo)
F_abs_cond   <- anova_abs_expo["condition.f", "F"]
ndf_abs_cond <- anova_abs_expo["condition.f", "num Df"]
ddf_abs_cond <- anova_abs_expo["condition.f", "den Df"]
p_abs_cond   <- anova_abs_expo["condition.f", "Pr(>F)"]

em_abs <- emmeans::emmeans(mod_returns_abs_expo, ~ condition.f,
                           pbkrtest.limit = 6000)
em_abs_df    <- as.data.frame(em_abs)
em_abs_ctrl  <- em_abs_df$emmean[em_abs_df$condition.f == "Control"]
em_abs_manip <- em_abs_df$emmean[em_abs_df$condition.f == "Manipulation"]
em_abs_ctrl_se  <- em_abs_df$SE[em_abs_df$condition.f == "Control"]
em_abs_manip_se <- em_abs_df$SE[em_abs_df$condition.f == "Manipulation"]
em_abs_diff <- em_abs_manip - em_abs_ctrl
```

So far we focused on analyzing behavior for the pre and post phases.
Here, we look at returns and investments in the exposure phase. The
linear mixed effects model of participants' returns in the exposure
phase does not show a main effect of Condition on returns. There was a
main effect of Investment,
`r papaja::apa_print(mod_returns_expo)$full_result$inv_scaled`, with
participants positively reciprocating higher investments, an interaction
effect between Condition and Investment
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2",papaja::apa_print(mod_returns_expo)$full_result$condition_f_inv_scaled)`,
showing a stronger positive reciprocity in the Control condition, and
finally a three way interaction between the RS group, Condition and
Investment
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2",papaja::apa_print(mod_returns_expo)$full_result$condition_f_inv_scaled_high_RS)`,
showing that this stronger positive reciprocity to investment in the
Control condition is higher for participants with high RS. The percentage-returns model's marginal and conditional $R^2$ were `r {.r <- performance::r2_nakagawa(mod_returns_expo$full_model); sprintf("%.3f and %.3f", .r$R2_marginal, .r$R2_conditional)}`, respectively. The linear
mixed effects model of the HMM investments shows a main effect of
Condition `r papaja::apa_print(mod_invs_expo)$full_result$condition_f`,
suggesting higher overall investments for the forgiving HMM compared to
the human-like HMM, but no difference in investments when facing low and
high RS groups.

In summary, despite the forgiving HMM sending overall higher investments
in the exposure phase, participants returned similar proportions of the
multiplied investments as those facing the human-like HMM. In absolute
terms, however, the higher investments combined with comparable per-unit
reciprocation produced higher returns in the Manipulation condition than
in the Control condition ($M$ = `r round(em_abs_manip, 2)` units vs.~$M$
= `r round(em_abs_ctrl, 2)` units; mixed-effects model
$F$(`r ndf_abs_cond`, `r round(ddf_abs_cond)`) =
`r round(F_abs_cond, 2)`, $p$ = `r papaja::printp(p_abs_cond)`).
The absolute-returns model's marginal and conditional $R^2$ were `r {.r <- performance::r2_nakagawa(mod_returns_abs_expo$full_model); sprintf("%.3f and %.3f", .r$R2_marginal, .r$R2_conditional)}`, respectively. Manipulation participants therefore did not behave more selfishly in
absolute terms during exposure; they matched the agent's higher
investments with proportional reciprocation. Beyond mean levels, the
slope of positive reciprocity of returns to investments was higher in
the Control condition, with this relationship stronger for the high
RS group.

### Questionnaire scores and performance

```{r}

temp2 <- d_finished %>% dplyr::filter(gameNum.f %in% c("pre", "post")) %>%
  dplyr::select(id, payoffTrust1, payoffTrust2, condition.f, RS_score, LPFS_score) %>%
  mutate(perf = payoffTrust1 + payoffTrust2) %>% 
  unique()
cor.test(temp2$LPFS_score, temp2$perf, method = "spearman")
cor.test(temp2$RS_score, temp2$perf, method = "spearman")

cor_result <- cor.test(temp2$RS_score, temp2$LPFS_score, method = "spearman")
# Format the result using papaja
apa_result <- papaja::apa_print(cor_result)

print(apa_result)
# Extract the correlation coefficient and p-value
spearman_rho <- apa_result$estimate

```

While we found a significant correlation between participants' Levels
of Personality Functioning Score (LPFS) and the Rejection Sensitivity
Questionnaire score (RSQ), Spearman's `r spearman_rho`, $p < .001$,
there was no correlation between these questionnaire scores and
participants' return or overall task performance.

<!---------------------- Modelling of co-player ratings -------------------------------------->

```{r afexModCoop, cache=TRUE}

mod_rating_coop <- mixed( rating_coop ~ gameNum.f*condition.f*high_RS + (1 | playerId), datRatings, REML= TRUE, method="KR")
saveRDS(summary(mod_rating_coop), "data/mod_rating_coop.RDS")

summary(mod_rating_coop)

```

```{r emmCoop, cache=TRUE}
# Contrast 1: Pre vs. Post
contrast1 <- c(-1, 0, 0, 0, 1)  

# Contrast 2: Pre vs. (Expo1, Expo2, Expo3) 
contrast2 <- c(-3, 1, 1, 1, 0) 

# Contrast 3: Post vs. (Expo1, Expo2, Expo3) 
contrast3 <- c(0, -1, -1, -1, 3) 

# Combine contrasts into a matrix
contrast_matrix <- rbind(contrast1, contrast2, contrast3)
colnames(contrast_matrix) <- levels(datRatings$gameNum.f) 


# Apply contrasts to the model using 'emmeans'
emm_coop <- emmeans(mod_rating_coop, ~ gameNum.f | condition.f)

# Custom comparisons: use 'contrast' directly
comp1_coop <- contrast(emm_coop, list("Pre vs Post" = contrast_matrix[1,]), by = "condition.f")
comp2_coop <- contrast(emm_coop, list("Pre vs Expo" = contrast_matrix[2,]), by = "condition.f")
comp3_coop <- contrast(emm_coop, list("Post vs Expo" = contrast_matrix[3,]), by = "condition.f")


comp1_coop

comp2_coop

comp3_coop

```

```{r}
# Compute the estimated marginal means (EMMs) for the interaction
emm_coop_interaction <- emmeans(mod_rating_coop, ~ gameNum.f | condition.f * high_RS)

comp1_coop_interaction <- contrast(emm_coop_interaction, list("PrePost" = contrast_matrix[1,]))
comp2_coop_interaction <- contrast(emm_coop_interaction, list("Pre vs Expo" = contrast_matrix[2,]))

# Extract summary for easier inline reporting
coop_summary <- summary(comp1_coop_interaction)
coop_results <- list(
  ControlHighRS = coop_summary[coop_summary$condition.f == "Control" & coop_summary$high_RS == "high RS", ],
  ControlLowRS = coop_summary[coop_summary$condition.f == "Control" & coop_summary$high_RS == "low RS", ],
  ManipHighRS = coop_summary[coop_summary$condition.f == "Manipulation" & coop_summary$high_RS == "high RS", ],
  ManipLowRS = coop_summary[coop_summary$condition.f == "Manipulation" & coop_summary$high_RS == "low RS", ]
)

# Output the results
print("Pre vs Post")
comp1_coop_interaction
```

```{r afexModForg}

mod_rating_forg <- mixed( rating_forgiving ~ gameNum.f*condition.f*high_RS + (1 | playerId), datRatings, REML= TRUE, method="KR")
saveRDS(summary(mod_rating_forg), "data/mod_rating_forg.RDS")

#summary(mod_rating_forg)

```

```{r emmForg}

# Apply contrasts to the model using 'emmeans'
emm_forg <- emmeans(mod_rating_forg, ~ gameNum.f | condition.f)

# Custom comparisons: use 'contrast' directly
comp1_forg <- contrast(emm_forg, list("Pre vs Post" = contrast_matrix[1,]), by = "condition.f")
comp2_forg <- contrast(emm_forg, list("Pre vs Expo" = contrast_matrix[2,]), by = "condition.f")
comp3_forg <- contrast(emm_forg, list("Post vs Expo" = contrast_matrix[3,]), by = "condition.f")


comp1_forg

comp2_forg

comp3_forg

```

```{r}
# Compute the estimated marginal means (EMMs) for the interaction
emm_forg_interaction <- emmeans(mod_rating_forg, ~ gameNum.f | condition.f * high_RS)

comp1_forg_interaction <- contrast(emm_forg_interaction, list("PrePost" = contrast_matrix[1,]))
comp2_forg_interaction <- contrast(emm_forg_interaction, list("Pre vs Expo" = contrast_matrix[2,]))

# Extract summary for easier inline reporting
forg_summary <- summary(comp1_forg_interaction)
forg_results <- list(
  ControlHighRS = forg_summary[forg_summary$condition.f == "Control" & forg_summary$high_RS == "high RS", ],
  ControlLowRS = forg_summary[forg_summary$condition.f == "Control" & forg_summary$high_RS == "low RS", ],
  ManipHighRS = forg_summary[forg_summary$condition.f == "Manipulation" & forg_summary$high_RS == "high RS", ],
  ManipLowRS = forg_summary[forg_summary$condition.f == "Manipulation" & forg_summary$high_RS == "low RS", ]
)

# Output the results
print("Pre vs Post")
comp1_forg_interaction
```

```{r afexModAgain}

mod_rating_again <- mixed( rating_playAgain ~ gameNum.f*condition.f*high_RS + (1 | playerId), datRatings, REML= TRUE, method="KR")
saveRDS(summary(mod_rating_again), "data/mod_rating_again.RDS")

#summary(mod_rating_again)

```

```{r emmAgain}

# Apply contrasts to the model using 'emmeans'
emm_again <- emmeans(mod_rating_again, ~ gameNum.f | condition.f)

# Custom comparisons: use 'contrast' directly
comp1_again <- contrast(emm_again, list("Pre vs Post" = contrast_matrix[1,]), by = "condition.f")
comp2_again <- contrast(emm_again, list("Pre vs Expo" = contrast_matrix[2,]), by = "condition.f")
comp3_again <- contrast(emm_again, list("Post vs Expo" = contrast_matrix[3,]), by = "condition.f")


comp1_again

comp2_again

comp3_again

```

```{r}
# Compute the estimated marginal means (EMMs) for the interaction
emm_again_interaction <- emmeans(mod_rating_again, ~ gameNum.f | condition.f * high_RS)

comp1_again_interaction <- contrast(emm_again_interaction, list("PrePost" = contrast_matrix[1,]))
comp2_again_interaction <- contrast(emm_again_interaction, list("Pre vs Expo" = contrast_matrix[2,]))

# Extract summary for easier inline reporting
again_summary <- summary(comp1_again_interaction)
again_results <- list(
  ControlHighRS = again_summary[again_summary$condition.f == "Control" & again_summary$high_RS == "high RS", ],
  ControlLowRS = again_summary[again_summary$condition.f == "Control" & again_summary$high_RS == "low RS", ],
  ManipHighRS = again_summary[again_summary$condition.f == "Manipulation" & again_summary$high_RS == "high RS", ],
  ManipLowRS = again_summary[again_summary$condition.f == "Manipulation" & again_summary$high_RS == "low RS", ]
)

# Output the results
print("Pre vs Post")
comp1_again_interaction
```

## Player ratings

Figure \@ref(fig:plotRatings) shows participants' ratings of co-players across phases. We examined two contrasts: pre-exposure versus exposure phase ratings, and pre-exposure versus post-exposure ratings.

High RS participants showed more differentiated perceptions of the agents. In the Manipulation condition, they rated the forgiving agents as more cooperative during exposure (`r papaja::apa_print(comp2_coop_interaction)$full_result$Manipulation_High_RS_Pre_vs_Expo`). In the Control condition, however, high RS participants rated the same human-like HMM progressively more negatively: lower on cooperation (`r papaja::apa_print(comp2_coop_interaction)$full_result$Control_High_RS_Pre_vs_Expo`), forgiveness (`r papaja::apa_print(comp2_forg_interaction)$full_result$Control_High_RS_Pre_vs_Expo`), and willingness to play again (`r papaja::apa_print(comp2_again_interaction)$full_result$Control_High_RS_Pre_vs_Expo`), despite the agent's strategy remaining unchanged. Low RS participants showed largely undifferentiated perceptions between pre and exposure phases regardless of condition.

Comparing pre to post-exposure ratings revealed a contrast effect: after experiencing the forgiving agent, both RS groups in the Manipulation condition rated the post-exposure agent (the same standard agent as in pre-exposure) more negatively on forgiveness (High RS: `r format_emmeans_apa(forg_results$ManipHighRS)`; Low RS: `r format_emmeans_apa(forg_results$ManipLowRS)`) and willingness to play again (High RS: `r format_emmeans_apa(again_results$ManipHighRS)`; Low RS: `r format_emmeans_apa(again_results$ManipLowRS)`). Low RS participants in the Control condition showed stable ratings, accurately perceiving the consistent agent strategy, while high RS participants in the Control condition continued their negative drift (Cooperation: `r format_emmeans_apa(coop_results$ControlHighRS)`). These rating patterns converge with the behavioral findings, suggesting high RS individuals are particularly sensitive to relative comparisons between interaction partners.

```{r plotRatings,include=T, echo=FALSE, fig.pos='H', fig.cap="Participants' ratings of co-players by phase, condition, and RS group. Blue: perceived cooperation; orange: perceived forgiveness; green: willingness to play again. Low RS participants showed stable ratings in the Control condition. High RS participants showed declining ratings in the Control condition despite unchanged agent strategy, and more differentiated perceptions in the Manipulation condition.", fig.align="center", fig.width=5, fig.height = 5}

ggplot(summary_df, aes(x = gameNum.f, y = mean, group = interaction(rating_type, condition.f), color = rating_type)) +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), width = .1, position = position_dodge(width = 0.25)) +
  geom_line(linetype = "dashed", position = position_dodge(width = 0.25)) +
  geom_point(position = position_dodge(width = 0.25), size = 3) +
  facet_wrap(~high_RS*condition.f, scales = "free_x") +
  labs(x = "Phase", y = "Mean Rating", color = "Rating Type") +
  scale_color_manual(values = c("rating_coop" = "#56B4E9", "rating_forgiving" = "#E69F00", "rating_playAgain" = "#009E73"),
                     labels = c("Cooperative", "Forgiving", "Play again")) +
  theme_bw() +
  theme(legend.position = "bottom")

```

```{r}
d_finished %>% 
  dplyr::select(id,Turing.choice) %>% 
  unique() %>% 
  group_by(Turing.choice) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)
```

<!-- Extracting Investor latent state from data -->

```{r extractInvStates}

library(tidyverse)

# Convert dataset from wide to long format, focusing on columns that start with "State_"
d_long <- d_finished %>%
  dplyr::select(id, condition.f, gameNum.f, gameNumber, roundNum, investorState.f) %>% 
  drop_na()
# View the resulting dataset
print(d_long)

```

```{r InvStatePlot}
library(ggplot2)
library(dplyr)

# Filter for only games 1 and 2
filtered_prepost <- d_long %>%
  filter(gameNum.f %in% c("pre", "post"))

# Define the colors for the levels of investorState

state_colors <- c("happy" = "Blue", "neutral" = "#77DD77", "unhappy" = "#FF6961")

# Create the histogram with stacked bars and specified colors
ggplot(filtered_prepost, aes(x = as.factor(roundNum), fill = investorState.f)) +
  geom_bar(aes(group = investorState.f), position = position_stack(reverse = FALSE), stat = "count") +
  scale_fill_manual(values = state_colors) +
  facet_wrap(~condition.f*gameNum.f, scales = "free_x", labeller = label_both) +
  labs(x = "Round", fill = "Investor State", title = "Histogram of Latent HMM Investor States per Game and Round") +
  theme_bw() +
  theme(legend.position = "bottom")

```

```{r}
# Filter for expo games only
filtered_expo <- d_long %>%
  filter(gameNum.f %in% c("expo1", "expo2", "expo3"))

# Create the histogram with stacked bars and specified colors
ggplot(filtered_expo, aes(x = as.factor(roundNum), fill = investorState.f)) +
  geom_bar(aes(group = investorState.f), position = position_stack(reverse = FALSE), stat = "count") +
  scale_fill_manual(values = state_colors) +
  facet_wrap(~condition.f*gameNum.f, scales = "free_x", labeller = label_both) +
  labs(x = "Round", fill = "Investor State", title = "Histogram of Latent States per Game and Round") +
  theme_bw() +
  theme(legend.position = "bottom")
```

<!------------------------ AVERAGING LATENT STATE COUNT ACROSS TRIALS ----------------------------->

```{r}
counts <- filtered_prepost %>%
  group_by(condition.f, gameNum.f, investorState.f,id) %>%
  summarise(count = n()) %>%
    summarise(count = sum(count))
  
# Calculate the average count per state per condition
total_rounds <- filtered_prepost  %>%
  group_by(condition.f, gameNum.f) %>%
  summarise(total_rounds = n()) 

state_summary  <- counts %>%
  left_join(total_rounds, by = c("condition.f", "gameNum.f")) %>% 
  mutate(state_percentage = count/total_rounds)

ggplot(state_summary , aes(x = gameNum.f, y = state_percentage, fill = investorState.f)) +
  geom_bar(stat = "identity") +
  facet_wrap(~condition.f) +
  labs(x = "Condition", y = "Proportion", fill = "Investor State") +
  theme_bw() +
  theme(legend.position = "bottom")

state_colors <- c("happy" = "Blue", "neutral" = "#77DD77", "unhappy" = "#FF6961")

# Create the histogram with stacked bars and specified colors
ggplot(state_summary , aes(x = gameNum.f, y = state_percentage, fill = investorState.f)) +
  geom_bar(aes(group = investorState.f), position = position_stack(reverse = FALSE), stat = "identity") +
  scale_fill_manual(values = state_colors) +
  facet_wrap(~condition.f*gameNum.f, scales = "free_x", labeller = label_both) +
  labs(x = "Round", fill = "Investor State", title = "Histogram of Latent States per Game and Round") +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r ratingsManipulationCheck, include=FALSE, cache=TRUE}
# -----------------------------------------------------------------------------
# Manipulation check on the exposure-phase ratings.
# Added in response to Reviewer 2, comment 6 (Collabra R&R, May 2026):
# tests whether the forgiving agent was subjectively rated as more positive
# than the standard agent during exposure (manipulation check on the ratings),
# and whether this between-condition difference varied by RS group.
# -----------------------------------------------------------------------------

expo_ratings <- datRatings %>% dplyr::filter(gameNum.f %in% c("expo1", "expo2", "expo3"))

mod_expo_coop  <- afex::mixed(rating_coop ~ condition.f * high_RS + (1 | playerId),
                              data = expo_ratings, REML = TRUE, method = "KR")
mod_expo_forg  <- afex::mixed(rating_forgiving ~ condition.f * high_RS + (1 | playerId),
                              data = expo_ratings, REML = TRUE, method = "KR")
mod_expo_again <- afex::mixed(rating_playAgain ~ condition.f * high_RS + (1 | playerId),
                              data = expo_ratings, REML = TRUE, method = "KR")
saveRDS(list(coop = summary(mod_expo_coop),
             forg = summary(mod_expo_forg),
             again = summary(mod_expo_again)),
        "data/mod_ratings_manipcheck.RDS")

a_coop  <- anova(mod_expo_coop)
a_forg  <- anova(mod_expo_forg)
a_again <- anova(mod_expo_again)

# Condition main effect: F, df, p
F_coop_cond  <- a_coop["condition.f", "F"];   p_coop_cond  <- a_coop["condition.f", "Pr(>F)"]
F_forg_cond  <- a_forg["condition.f", "F"];   p_forg_cond  <- a_forg["condition.f", "Pr(>F)"]
F_again_cond <- a_again["condition.f", "F"];  p_again_cond <- a_again["condition.f", "Pr(>F)"]
# Condition x RS interaction: F, df, p
F_coop_int   <- a_coop["condition.f:high_RS", "F"];   p_coop_int   <- a_coop["condition.f:high_RS", "Pr(>F)"]
F_forg_int   <- a_forg["condition.f:high_RS", "F"];   p_forg_int   <- a_forg["condition.f:high_RS", "Pr(>F)"]
F_again_int  <- a_again["condition.f:high_RS", "F"];  p_again_int  <- a_again["condition.f:high_RS", "Pr(>F)"]
# Denominator df (same for all three: 202)
ddf_manip <- a_coop["condition.f", "den Df"]

# EMM by Condition (averaged over RS) for each rating
em_coop  <- as.data.frame(emmeans::emmeans(mod_expo_coop,  ~ condition.f, pbkrtest.limit = 5000))
em_forg  <- as.data.frame(emmeans::emmeans(mod_expo_forg,  ~ condition.f, pbkrtest.limit = 5000))
em_again <- as.data.frame(emmeans::emmeans(mod_expo_again, ~ condition.f, pbkrtest.limit = 5000))
coop_M_ctrl   <- em_coop$emmean[em_coop$condition.f == "Control"]
coop_M_manip  <- em_coop$emmean[em_coop$condition.f == "Manipulation"]
forg_M_ctrl   <- em_forg$emmean[em_forg$condition.f == "Control"]
forg_M_manip  <- em_forg$emmean[em_forg$condition.f == "Manipulation"]
again_M_ctrl  <- em_again$emmean[em_again$condition.f == "Control"]
again_M_manip <- em_again$emmean[em_again$condition.f == "Manipulation"]
```

Focusing specifically on the exposure phase, we used the ratings as a
manipulation check to test whether the forgiving agent produced a more
positive subjective experience than the standard agent, and whether
this differed by RS group. Each of the
three exposure-phase ratings was modeled as a function of Condition
$\times$ RS with player random intercepts. Manipulation participants
rated the agent as more enjoyable to play with again ($F$(1,
`r round(ddf_manip)`) = `r sprintf("%.2f", F_again_cond)`, $p$ =
`r papaja::printp(p_again_cond)`), and marginally higher on
cooperativeness ($F$(1, `r round(ddf_manip)`) =
`r sprintf("%.2f", F_coop_cond)`, $p$ = `r papaja::printp(p_coop_cond)`)
and forgivingness ($F$(1, `r round(ddf_manip)`) =
`r sprintf("%.2f", F_forg_cond)`, $p$ = `r papaja::printp(p_forg_cond)`).
The Condition $\times$ RS interaction was not significant on any of the
three ratings (cooperative: $F$(1, `r round(ddf_manip)`) =
`r sprintf("%.2f", F_coop_int)`, $p$ = `r papaja::printp(p_coop_int)`;
forgiving: $F$(1, `r round(ddf_manip)`) =
`r sprintf("%.2f", F_forg_int)`, $p$ = `r papaja::printp(p_forg_int)`;
play again: $F$(1, `r round(ddf_manip)`) =
`r sprintf("%.2f", F_again_int)`, $p$ = `r papaja::printp(p_again_int)`),
indicating no evidence that the manipulation was registered differently
by high-RS and low-RS participants.


# Discussion

We used a randomized controlled online experiment where participants
played a RTG with artificial agents designed to simulate human-like
trust-building scenarios. Participants were then exposed to either
forgiving and behaviorally consistent HMM agents (which, by design, were also more cooperative) or standard human-like HMM agents before playing another RTG. We found
that RS did not moderate the overall effect of exposure on
participants' returns as trustees in the repeated trust game. While previous research has shown that RS affects
*trust* formation, appraisal and repair, its impact on *reciprocity* in
repeated economic exchanges has been less explored. Our results suggest
a potential dissociation between RS's well-documented effects on broader social
behavior and its limited influence on reciprocity in structured,
repeated interactions. This specificity also contrasts with recent
work on malevolent traits, in which trustees selected for a high dark
factor of personality returned systematically lower proportions when
facing comparable HMM investors [@guennouni_price_2026], whereas RS
altered when and how cooperation recovered rather than how much
participants returned overall.

The design deliberately created two qualitatively different exposure experiences: Manipulation participants interacted with partners who were both forgiving (unable to remain in low-trust states) *and* behaviorally consistent (no pre-programmed trust withdrawals), while Control participants interacted with partners who maintained the same pattern of occasional low investments seen in the pre-exposure phase. This design reflects the multidimensional nature of secure relational experiences in attachment theory: a secure base provides both responsiveness to distress (forgiveness) and consistent availability [@bowlby_secure_1988]. The Control condition thus represents continuity with typical relationship patterns, while the Manipulation condition tests whether exposure to an idealized partner (one who is positive on both dimensions) produces lasting change.

Contrary to our hypothesis, exposure to forgiving and behaviorally
consistent agents did not increase participants' reciprocity or
cooperation, nor did it prompt the artificial agent to increase its
trust in participants through higher investments. Instead, participants reduced their returns overall while
the returns of those in the Control group did not change between the pre
and post phase of the experiment. Why did participants reduce their
returns even though they were repeatedly exposed to agents designed to be more forgiving? How participants rated their co-players helps explain this reduction. Those
exposed to the forgiving agent rated their opponent in the post-exposure
phase lower on all attributes even though they faced the same dynamic
human-like HMM as pre-exposure. One possible explanation for this drop
in rating is a negative contrast effect, whereby evaluation of a person
or situation is skewed by comparison with a recently encountered, more
favorable one, so that ordinary behavior seems less favorable
[@kobre_negative_1972]. As the most recently faced opponents
were more forgiving (and consequently more cooperative), this negative contrast effect may
have trumped any learning transfer from being repeatedly exposed to
forgiving agents [@zentall_within-trial_2005]. This pattern is
conceptually related to findings from norm-training paradigms in
economic exchange, where pre-adaptation to different partner profiles
produces contrast effects on subsequent judgments of unrelated partners
[@xiang_computational_2013]. Our manipulation differs from
norm-training in *what was manipulated*: rather than exposing
participants to offers drawn from a shifted level distribution
[@xiang_computational_2013], we changed the partner's *response dynamics*
(the transition function governing how the forgiving agent recovered
from low-trust states). Because the forgiving agent could not dwell in
low trust, these dynamics also raised its realized investments, so the
participant's reference for a "normal" partner shifted along both
responsiveness and investment level.

The post-exposure reduction invites explanations beyond negative contrast, and our data help discriminate among them. Betrayal aversion (heightened sensitivity to the post-phase low investment, given that exposure contained no withdrawals) predicts group differences only after that low investment. Instead, Manipulation participants had already reduced their returns in rounds 1-12 of the post-exposure phase, before any low investment occurred, consistent with contrast operating from the first post-exposure round. Betrayal aversion may contribute to specific patterns, such as the delayed recovery of high RS participants, but cannot explain a reduction that preceded any withdrawal.

A second account is strategic exploitation: a less punitive partner might teach participants they can return less without lasting consequence [@thielmann_personality_2020]. This predicts more investor-harming returns and higher trustee payoffs in the Manipulation condition; instead, Manipulation participants showed lower exploitation rates than Control at both phases, no Phase × Condition interaction, and no rise in trustee payoffs. They were not extracting more from their partners.

A subtler variant of the strategic account is contingency-based adaptation: that participants returned less because, under the forgiving agent, low returns carried weaker consequences for future investment [@balliet_reward_2011]. This does not fit the agent's design: as described in the Methods, the forgiving agent still punished low returns immediately, at the same rate as the standard agent, differing only in not sustaining that punishment, so a contingency account would require participants to track punishment persistence over a multi-round horizon, a demanding inference given the modest bonus. The data also weigh against it directly: behavior during exposure itself showed no such adaptation (Manipulation participants returned the same proportions as Control, and more in absolute terms, while the weakened contingencies were actually in force); the post-exposure agent followed the same policy as before, yet participants rated it more negatively after exposure (notably on forgiveness and willingness to play again), a perception shift toward an unchanged partner that negative contrast predicts but a contingency account does not; and the reduction preceded any post-phase low investment. Finally, any uniform mechanism predicts a same-direction change for all Manipulation participants, whereas the four-way Phase × Condition × Investment × RS interaction shows a reversal by RS group (high RS less responsive, low RS more responsive). Because this interaction was planned but not separately powered, we treat it as corroborating rather than decisive, and regard contingency-based adaptation as a plausible partial contributor, most likely in the high RS subgroup, rather than the primary driver.

These results also bear on what motivates trustworthiness more generally. Three accounts dominate, each tying the trustee's return to a different feature of the partner: reciprocity preferences to the perceived kindness of its investment [@falk_theory_2006], inequity aversion to the resulting payoff difference [@fehr_theory_1999], and vulnerability-responsiveness to the vulnerability the trustor exposes [@cox_what_2016]. Because the post-exposure agent followed the same policy as before, all three predict unchanged returns; each can accommodate the reduction only by adding a shift in the trustee's reference point, which is the negative contrast mechanism we propose. The reduction therefore implicates a reference-dependent process rather than any of these preferences acting alone.

While the negative contrast effect operated across RS groups, the pattern of responses to trust withdrawals differed by RS in ways that parallel clinical models of rejection sensitivity. High and low RS participants showed comparable immediate reactions to the low investment (the Drop), indicating intact detection of trust withdrawals regardless of RS level; the groups diverged only in the timing of recovery, with high RS participants in the Manipulation condition recovering later than low RS and Control participants (the between-condition gap closed by the second post-breach round). This dissociation between intact detection and a brief delay in re-establishing cooperation is reminiscent of social-learning difficulties reported in elevated RS and related presentations: studies of borderline personality disorder document slower incorporation of positive partner behavior into trust-related expectations [@king-casas_rupture_2008; @korn_social_2016], and work on depression describes "cognitive immunization" against belief-disconfirming evidence [@kube_distorted_2020]. We advance these as analogies rather than equivalences: the RS effects here rest on planned but not separately powered interactions in a non-clinical extreme-groups sample and require confirmatory replication before any clinical interpretation.

The four-way interaction is consistent with this interpretation, though it is exploratory. High RS participants showed reduced responsiveness to their co-player's investments following the manipulation, a pattern that parallels the social withdrawal documented in high RS populations, where anticipatory self-protection can undermine relationship maintenance [@ayduk_rejection_2000; @romero-canyas_rejection_2010]. Low RS participants instead showed increased responsiveness post-exposure, consistent with having internalized the cooperative behavior experienced during exposure. This differential pattern aligns with broader findings that RS is associated with difficulty deploying adaptive interpersonal strategies [@ayduk_regulating_2000].

Together, these signatures (reduced responsiveness and a brief recovery delay) suggest a pattern of passive disengagement rather than active retaliation: RS modulated specific aspects of cooperative behavior without shifting overall return levels. The ratings data complement these findings: high RS participants showed more negative explicit evaluations of their co-players, rating them lower on forgiveness and willingness to interact again. This convergence suggests RS effects are expressed across multiple response systems: the structured trust game may confine them to specific behavioral signatures, whereas the open-ended ratings allow broader expression of RS-related evaluative biases [@lieberman_social_2007].

These findings have implications for interventions aimed at promoting trust and cooperation. The present results suggest that exposure to forgiving partner behavior alone may be insufficient for high RS individuals, and may even be counterproductive through negative contrast. The specific patterns observed, a brief delay in recovery from trust withdrawals and reduced sensitivity to partner behavior, point to potential intervention targets. Approaches that focus on enhancing the capacity to update expectations following interpersonal ruptures may be more effective than relying on this kind of exposure alone. Future research should examine whether these behavioral patterns generalize to naturalistic social contexts and whether targeted interventions can modify the updating and recovery patterns observed in high RS participants. However, given the minimal nature of the trust game paradigm and the brief exposure period, these findings should be interpreted as proof-of-concept demonstrations rather than direct evidence for clinical intervention design. The corrective experience hypothesis in attachment theory typically refers to sustained, emotionally significant relationships, so whether the effects of brief exposure scale to therapeutic contexts remains an open question.

## Limitations

While this study clarifies how exposure to forgiving partners shapes
subsequent cooperation, several limitations warrant consideration. First, as described in the Methods, the Manipulation condition combined forgiveness with the absence of programmed trust withdrawals; this means we tested a "strong" version of the corrective experience hypothesis rather than isolating forgiveness specifically, whereas Control participants experienced realistic continuity. Future replications could include additional conditions (e.g., forgiving agents that still deliver occasional low investments) to isolate the unique contribution of each relational dimension. Relatedly, because the forgiving agent's higher realized investments were themselves a consequence of these dynamics, the present design cannot separate participants' responses to the recovery dynamics from their responses to the higher investment level those dynamics produced. Second, our extreme groups design for RS (selecting
participants with RSQ scores > 15 or < 10) was chosen to maximize
sensitivity to RS moderation but may inflate effect sizes and limit
generalizability to individuals with moderate RS. Future research should
examine RS as a continuous variable. Third, the brief exposure phase
(three seven-round games) may have been insufficient to induce lasting
changes. Fourth, the online format eliminates social cues present in
face-to-face interactions. Notably, although `r pct_machine`% of
participants correctly identified the agent as a machine, the
post-exposure reduction in returns held both among those who detected
the agent and among the `r pct_fooled`% who did not (formal sensitivity
analysis reported above in "Sensitivity to suspicion of AI agent"),
suggesting the effect does not depend on the cover story being believed. Despite these limitations, subsequent studies could
address these constraints by incorporating face-to-face interactions,
longer exposure periods, and continuous RS measurement. Regarding
generalizability, our results may be specific to adults with high or low
RS recruited from online platforms and may not generalize to clinical
populations, children, or older adults. The core finding of
decreased cooperation after exposure to forgiving agents may generalize
across populations and contexts, but its strength, its interaction with
RS, and its sensitivity to cultural norms of cooperation and trust
remain open questions for cross-cultural replication.

# Conclusion

This randomized controlled experiment uncovered an unexpected effect of
exposure to forgiving agents: rather than promoting cooperation, it
reduced subsequent returns through a negative contrast effect, while
rejection sensitivity shaped specific behavioral signatures rather than
overall cooperation levels. These findings challenge existing
assumptions about fostering cooperative behavior and suggest the need
for more nuanced interventions. Importantly, the use of HMM-based
artificial agents in this study represents a significant methodological
advancement. By providing a balance between experimental control and
realistic, adaptive behavior, these agents allowed for a nuanced
exploration of trust dynamics that would be challenging to achieve with
human confederates or simplistic computer algorithms. This approach
opens up new possibilities for studying complex social interactions in
controlled settings, potentially bridging the gap between laboratory
experiments and real-world social dynamics.

\pagebreak

# Author contributions statement {.unnumbered}

Author A, Author B and Author C designed and developed the study
concept. Experiment design, testing and data collection were performed
by Author A. Author A analyzed and interpreted the data under
the supervision of Author B and Author C. All authors jointly wrote and
approved the final version of the manuscript for submission.



# Funding {.unnumbered}

[Anonymized for double-blind review.]



# Competing interests statement {.unnumbered}

The author(s) declared that there were no conflicts of interest with
respect to the authorship or the publication of this article.

# Acknowledgements {.unnumbered}

[Anonymized for double-blind review.]



# Additional information {.unnumbered}

## Correspondence {.unnumbered}

All correspondence and requests for materials should be addressed to
the corresponding author.



## Transparency and data availability {.unnumbered}

Preregistration: The hypotheses and methods were not preregistered. The primary hypothesis, based on attachment theory, predicted that forgiveness exposure would increase cooperation. The alternative predictions regarding contrast effects and RS-specific updating biases were incorporated into the theoretical framework following initial data analysis, though both perspectives were grounded in existing literature. The analysis plan was not preregistered. Materials, data, and analysis
scripts: All study materials, primary data, and analysis scripts are
available in an anonymized repository for review at <https://github.com/jjg6wv74hb-stack/forgiveness-trust-rs>. Upon publication, all materials, data, and analysis code will be deposited in a public repository with a persistent identifier (DOI).



# References
